---
title: "Story 3: Do Stricter Gun Laws Reduce Firearm Mortality?"
author: "Randy Howk"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
    theme: readable
---

# Overview

This story explores whether **stricter firearm control laws** are associated with **lower firearm mortality** (deaths per 100,000 persons), using:

- **CDC injury/violence mapping datasets** accessed via API (Socrata / data.cdc.gov)
- A **5-point Likert** scale of gun-law strictness derived from **GIFFORDS state grades**
- State-level CDC mortality plus GIFFORDS strictness grading

> **Important note (historical context):** The CDC’s *mortality data collection* (via the National Vital Statistics System) continues across decades; the major policy shift in the mid-1990s primarily reduced/removed dedicated federal funding for firearm injury research rather than stopping mortality reporting. We include a long-run time window beginning in **1999** (consistent annual coverage in these CDC mapping datasets), and you can extend/adjust the window as needed.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
  library(glue)
  library(scales)
  library(usmap)
  library(viridis)
  library(rvest)
  library(officer)
  library(rvg)
})

# Guardrail for IDE runs: render this Rmd, not CSV debug artifacts.
current_input <- tryCatch(basename(knitr::current_input()), error = function(e) "")
if (nzchar(current_input) && current_input != "Story3.rmd") {
  message(
    "This document is intended to run from Story3.rmd. ",
    "Do not render CSV debug files directly."
  )
}

# Normalize state labels from mixed sources to 2-letter USPS abbreviations.
# Handles full state names, plain abbreviations, and labels like "Alabama (AL)".
to_state_abbr <- function(x) {
  x <- as.character(x)
  x <- stringr::str_squish(x)
  x <- stringr::str_to_title(x)

  # If already a plain 2-letter code, keep it.
  is_code <- stringr::str_detect(x, "^[A-Za-z]{2}$")
  out <- rep(NA_character_, length(x))
  out[is_code] <- stringr::str_to_upper(x[is_code])

  # Parse labels with trailing "(XX)" code.
  paren_code <- stringr::str_match(x, "\\(([A-Za-z]{2})\\)$")[, 2]
  out[is.na(out) & !is.na(paren_code)] <- stringr::str_to_upper(paren_code[is.na(out) & !is.na(paren_code)])

  # Strip any parenthetical suffix and map by full name.
  base_name <- stringr::str_squish(stringr::str_replace(x, "\\s*\\(.*\\)$", ""))
  idx <- match(base_name, state.name)
  out[is.na(out) & !is.na(idx)] <- state.abb[idx[is.na(out) & !is.na(idx)]]

  # Keep only true US states for this project.
  out[!out %in% state.abb] <- NA_character_
  out
}
```

# Data sources

## CDC firearm mortality (API)

We use CDC “Mapping Injury, Overdose, and Violence” datasets hosted on **data.cdc.gov**:

- **State dataset:** `fpsi-y8tj`
- **County dataset:** `psx4-wq38`

These datasets can be accessed using Socrata’s SODA API (CSV endpoint).

```{r cdc-pull}
# SODA CSV endpoints (large limits; adjust if you hit size constraints)
cdc_state_url  <- "https://data.cdc.gov/resource/fpsi-y8tj.csv?$limit=500000"
cdc_county_url <- "https://data.cdc.gov/resource/psx4-wq38.csv?$limit=500000"

cdc_state_raw  <- readr::read_csv(cdc_state_url, show_col_types = FALSE) |> clean_names()
cdc_county_raw <- readr::read_csv(cdc_county_url, show_col_types = FALSE) |> clean_names()

glimpse(cdc_state_raw)
glimpse(cdc_county_raw)

message("CDC state raw rows: ", nrow(cdc_state_raw), " | cols: ", ncol(cdc_state_raw))
message("CDC county raw rows: ", nrow(cdc_county_raw), " | cols: ", ncol(cdc_county_raw))

# Write small raw snapshots for debugging when downstream filters go empty.
readr::write_csv(utils::head(cdc_state_raw, 200), "cdc_state_raw_sample.csv")
readr::write_csv(utils::head(cdc_county_raw, 200), "cdc_county_raw_sample.csv")
```

## Filter to firearm mortality & years

```{r cdc-filter}
start_year <- 2017

state_intents <- sort(unique(as.character(cdc_state_raw$intent)))
county_intents <- sort(unique(as.character(cdc_county_raw$intent)))

message("Sample CDC state intents: ", paste(head(state_intents, 10), collapse = ", "))
message("Sample CDC county intents: ", paste(head(county_intents, 10), collapse = ", "))

# Prefer FA_Deaths when available; otherwise fall back to legacy firearm labels.
state_keep <- intersect(c("FA_Deaths"), state_intents)
county_keep <- intersect(c("FA_Deaths"), county_intents)

if (length(state_keep) > 0) {
  cdc_state <- cdc_state_raw |>
    filter(intent %in% state_keep)
  message("CDC state intent basis: FA_Deaths")
} else {
  cdc_state <- cdc_state_raw |>
    filter(str_detect(intent, regex("firearm", ignore_case = TRUE)))
  message("CDC state fallback basis: legacy /firearm/i intents")
}

if (length(county_keep) > 0) {
  cdc_county <- cdc_county_raw |>
    filter(intent %in% county_keep)
  message("CDC county intent basis: FA_Deaths")
} else {
  cdc_county <- cdc_county_raw |>
    filter(str_detect(intent, regex("firearm", ignore_case = TRUE)))
  message("CDC county fallback basis: legacy /firearm/i intents")
}

cdc_state <- cdc_state |>
  mutate(period = suppressWarnings(as.integer(period))) |>
  filter(!is.na(period), period >= start_year)

cdc_county <- cdc_county |>
  mutate(period = suppressWarnings(as.integer(period))) |>
  filter(!is.na(period), period >= start_year)

message("CDC state rows after firearm/year filter: ", nrow(cdc_state))
message("CDC county rows after firearm/year filter: ", nrow(cdc_county))

if (nrow(cdc_state) == 0) {
  readr::write_csv(utils::head(cdc_state_raw, 500), "cdc_state_filter_debug.csv")
  stop(
    "CDC state data is empty after filter. ",
    "Check cdc_state_raw_sample.csv and cdc_state_filter_debug.csv ",
    "for available columns/values (especially `intent`)."
  )
}

range(cdc_state$period, na.rm = TRUE)
```


# Gun law strictness: 5-point Likert (GIFFORDS grades)

## Methods and rationale for scraping

GIFFORDS publishes an annual state **Gun Law Scorecard** that assigns each state a **letter grade** (A+ … F) based on the strength of its firearm laws. These grades are widely cited, but GIFFORDS does **not** provide a stable, official multi-year machine-readable download suitable for reproducible analysis. The most reliable way to construct a multi-year panel (within the scope of this course project) is therefore to **programmatically extract** the state grades from GIFFORDS’ publicly available annual scorecard pages.

To keep the workflow reproducible and respectful:

- We **cache** the extracted results to a local CSV (`giffords_grades_2012_2024.csv`) so repeated knitting does not repeatedly hit the site.
- We include a **scrape “health check”** to ensure each year returns ~50 states; if a year fails, we stop with a clear error.
- We normalize away GIFFORDS’ on-page link tokens (e.g., `【162†Colorado】`) that can otherwise break simple text parsing.
- If GIFFORDS changes page structure, you can delete the cached CSV and re-knit to regenerate, or adjust the parsing function below.

### Likert mapping

We collapse GIFFORDS letter grades to a 5-point Likert scale:

- **5** = A+, A, A-
- **4** = B+, B, B-
- **3** = C+, C, C-
- **2** = D+, D, D-
- **1** = F

## Scrape and build a multi-year grade panel (2012–2024)

```{r giffords-scrape-multiyear}
scorecard_url <- function(year) {
  if (year == 2024) {
    "https://giffords.org/resources/scorecard2024/"
  } else {
    paste0("https://giffords.org/lawcenter/resources/scorecard", year, "/")
  }
}

grade_to_likert <- function(g){
  g <- stringr::str_trim(g)
  dplyr::case_when(
    stringr::str_detect(g, "^A") ~ 5,
    stringr::str_detect(g, "^B") ~ 4,
    stringr::str_detect(g, "^C") ~ 3,
    stringr::str_detect(g, "^D") ~ 2,
    stringr::str_detect(g, "^F") ~ 1,
    TRUE ~ NA_real_
  )
}

scrape_scorecard_year <- function(year) {
  url <- scorecard_url(year)
  message("Scraping GIFFORDS year ", year, " -> ", url)

  page <- rvest::read_html(url)

  # 1) Prefer real HTML tables if present
  tabs <- page |> rvest::html_elements("table") |> rvest::html_table(fill = TRUE)

  if (length(tabs) > 0) {
    cand <- tabs[[which.max(purrr::map_int(tabs, nrow))]] |> janitor::clean_names()

    if (any(stringr::str_detect(names(cand), "grade")) &&
        any(stringr::str_detect(names(cand), "state|name"))) {

      state_col <- names(cand)[which(stringr::str_detect(names(cand), "state|name"))[1]]
      grade_col <- names(cand)[which(stringr::str_detect(names(cand), "grade"))[1]]

      out <- cand |>
        dplyr::transmute(
          state = as.character(.data[[state_col]]) |> stringr::str_squish(),
          grade = as.character(.data[[grade_col]]) |> stringr::str_squish(),
          year  = year,
          likert = grade_to_likert(grade)
        ) |>
        dplyr::filter(state %in% state.name) |>
        dplyr::distinct()

      if (nrow(out) >= 45) return(out)
    }
  }

  # 2) Fallback: parse the page text (robust to the "【id†State】" link tokens)
  txt <- page |> rvest::html_text2()

  marker <- "Gun Law Strength"
  pos <- stringr::str_locate(txt, stringr::fixed(marker, ignore_case = TRUE))[1,1]
  if (is.na(pos)) stop("Could not locate the scorecard table text for year ", year)

  tail_txt <- stringr::str_sub(txt, pos, pos + 30000)

  # Remove the embedded "【...】" tokens that otherwise glue state names to grades
  tail_txt <- stringr::str_replace_all(tail_txt, "【[^】]*】", " ")
  tail_txt <- stringr::str_squish(tail_txt)

  m <- stringr::str_match_all(
    tail_txt,
    "\\b\\d{1,2}\\s+([A-Za-z ]+?)\\s+(A\\+|A-|A|B\\+|B-|B|C\\+|C-|C|D\\+|D-|D|F)\\b"
  )[[1]]

  if (nrow(m) == 0) stop("No (state, grade) pairs found for year ", year)

  tibble::tibble(
    state = stringr::str_squish(m[,2]),
    grade = m[,3],
    year  = year
  ) |>
    dplyr::filter(state %in% state.name) |>
    dplyr::mutate(likert = grade_to_likert(grade)) |>
    dplyr::distinct()
}

cache_file <- "giffords_grades_2012_2024.csv"

if (file.exists(cache_file)) {
  giffords_grades <- readr::read_csv(cache_file, show_col_types = FALSE)
} else {
  years <- 2012:2024
  giffords_grades <- purrr::map_dfr(years, ~{
    tryCatch(scrape_scorecard_year(.x),
             error = function(e){
               warning("Failed year ", .x, ": ", conditionMessage(e))
               tibble::tibble(state = character(), grade = character(), year = integer(), likert = numeric())
             })
  }) |>
    dplyr::arrange(year, state)

  readr::write_csv(giffords_grades, cache_file)
}

# Align both sources to project window: 2017 through latest available.
giffords_grades <- giffords_grades |>
  dplyr::mutate(year = suppressWarnings(as.integer(year))) |>
  dplyr::filter(!is.na(year), year >= start_year)

# Scrape health check: each year should have (almost) all 50 states
year_counts <- giffords_grades |> dplyr::count(year, name = "n_states")
print(year_counts, n = Inf)

cat("\n--- GIFFORDS SAMPLE ---\n")

print(
  giffords_grades %>%
    select(state, year, grade) %>%
    head(15)
)

cat("\nGIFFORDS rows per year:\n")
print(giffords_grades %>% count(year))

cat("\nUnique GIFFORDS states:\n")
print(sort(unique(giffords_grades$state)))

bad_years <- year_counts |> dplyr::filter(n_states < 45)

if (nrow(bad_years) > 0) {
  warning(
    "Some GIFFORDS years returned fewer than 45 states: ",
    paste(bad_years$year, collapse = ", ")
  )
}

# Choose a single analysis year using overlap + CDC coverage.
# This avoids selecting a very recent partial CDC year with too few states.
cdc_debug <- cdc_state |>
  dplyr::transmute(
    period_raw = as.character(period),
    period = suppressWarnings(as.integer(period)),
    name = as.character(name),
    state_abbr = to_state_abbr(name)
  ) |>
  dplyr::filter(!is.na(period), period >= start_year)

giffords_debug <- giffords_grades |>
  dplyr::transmute(
    state = as.character(state),
    year = suppressWarnings(as.integer(year)),
    grade = as.character(grade),
    likert = as.numeric(likert),
    state_abbr = to_state_abbr(state)
  ) |>
  dplyr::filter(!is.na(year), year >= start_year)

# Debug exports to inspect matching/years when knitting fails
# Note: these are diagnostics only. Render Story3.rmd, not cdc.csv/giffords.csv.
readr::write_csv(cdc_debug, "cdc.csv")
readr::write_csv(giffords_debug, "giffords.csv")

cdc_year_coverage <- cdc_debug |>
  dplyr::select(period, state_abbr) |>
  dplyr::filter(!is.na(period), !is.na(state_abbr)) |>
  dplyr::distinct(period, state_abbr) |>
  dplyr::count(period, name = "n_states") |>
  dplyr::arrange(desc(period))

cat("\nCDC state coverage by year (top 10):\n")
print(head(cdc_year_coverage, 10))

giffords_years <- sort(unique(giffords_debug$year[!is.na(giffords_debug$year)]))
common_years <- sort(intersect(giffords_years, cdc_year_coverage$period))

if (length(common_years) > 0) {
  expected_years <- seq(start_year, max(common_years), by = 1)
  missing_years <- setdiff(expected_years, common_years)
  if (length(missing_years) > 0) {
    warning("Missing overlap years in 2017+ window: ", paste(missing_years, collapse = ", "))
  }
}

overlap_debug <- tibble::tibble(
  cdc_years = paste(sort(unique(cdc_year_coverage$period)), collapse = ";"),
  giffords_years = paste(sort(unique(giffords_years)), collapse = ";"),
  common_years = paste(common_years, collapse = ";")
)
readr::write_csv(overlap_debug, "overlap_debug.csv")

if (length(common_years) == 0) {
  stop(
    "No overlapping year found between CDC and GIFFORDS. ",
    "Wrote debug files: cdc.csv, giffords.csv, overlap_debug.csv"
  )
}

year_quality <- tibble::tibble(year = common_years) |>
  dplyr::left_join(
    cdc_year_coverage |>
      dplyr::transmute(year = period, cdc_states = n_states),
    by = "year"
  ) |>
  dplyr::mutate(score = cdc_states)

cat("\nOverlapping CDC/GIFFORDS years:\n")
print(year_quality |> dplyr::arrange(desc(year)))

law_year_used <- year_quality |>
  dplyr::arrange(desc(score), desc(year)) |>
  dplyr::slice(1) |>
  dplyr::pull(year)
gun_law <- giffords_grades |>
  dplyr::filter(year == law_year_used) |>
  dplyr::transmute(
    state = stringr::str_to_title(stringr::str_squish(as.character(state))),
    grade = as.character(grade) |> stringr::str_squish(),
    likert = as.numeric(likert),
    state_abbr = to_state_abbr(state)
  ) |>
  dplyr::filter(!is.na(state_abbr))


message("Using GIFFORDS law-year: ", law_year_used)
```


# Analysis dataset

We analyze the **latest available year** for a cross-sectional comparison, and also build a long-run panel for optional trends.

```{r analysis-datasets}
# Align the analysis year so the law-grade year matches the CDC mortality year used
analysis_year <- law_year_used

state_latest <- cdc_state |>
  filter(period == analysis_year) |>
  transmute(
    state = stringr::str_to_title(stringr::str_squish(as.character(name))),
    rate  = as.numeric(rate),
    state_abbr = to_state_abbr(name)
  ) |>
  filter(!is.na(state_abbr), !is.na(rate)) |>
  group_by(state_abbr) |>
  summarize(
    state = first(state),
    # Single-intent FA_Deaths rate (or fallback proxy where unavailable).
    rate = mean(rate, na.rm = TRUE),
    .groups = "drop"
  )

cat("\n--- CDC SAMPLE ---\n")

print(
  state_latest %>%
    select(state, rate) %>%
    head(15)
)

cat("\nUnique CDC states:\n")
print(sort(unique(state_latest$state)))

analysis_df <- state_latest |>
  left_join(gun_law, by = "state_abbr", suffix = c("_cdc","_giff")) |>
  transmute(
    state = state_cdc,
    state_abbr = state_abbr,
    rate_per_100k = rate,
    grade = grade,
    likert = likert
  ) |>
  drop_na(rate_per_100k, likert) |>
  mutate(
    rate_per_capita = rate_per_100k / 100000,
    law_band = factor(likert, levels = 1:5, labels = c("F", "D", "C", "B", "A"), ordered = TRUE),
    strictness_group = dplyr::case_when(
      likert %in% c(4, 5) ~ "Strict (4-5)",
      likert %in% c(1, 2, 3) ~ "Less strict (1-3)",
      TRUE ~ NA_character_
    )
  ) |>
  filter(state_abbr != "PR")


# Sanity checks (helps prevent silent empty joins)
message("Rows in CDC state_latest: ", nrow(state_latest))
message("Rows in GIFFORDS gun_law: ", nrow(gun_law))
message("Rows after join (analysis_df): ", nrow(analysis_df))

if (nrow(analysis_df) < 10) {
  message("Top unmatched CDC states (not found in GIFFORDS for this year):")
  print(dplyr::anti_join(state_latest, gun_law, by = "state_abbr") |> dplyr::slice_head(n = 15))
  stop("analysis_df is too small after joining CDC + GIFFORDS. Check scraping output / state-name normalization.")
}

analysis_df |> arrange(desc(rate_per_100k)) |> head(10)
```

# Do stricter laws reduce firearm mortality?

## Summary by Likert bin

```{r by-likert}
by_likert <- analysis_df |>
  group_by(likert) |>
  summarize(
    states = n(),
    mean_rate_per_100k = mean(rate_per_100k, na.rm = TRUE),
    median_rate_per_100k = median(rate_per_100k, na.rm = TRUE),
    mean_rate_per_capita = mean(rate_per_capita, na.rm = TRUE),
    median_rate_per_capita = median(rate_per_capita, na.rm = TRUE)
  ) |>
  arrange(likert)

by_likert
```

## Combined groups comparison: strict (4–5) vs less strict (1–3)

```{r strict-vs-less}
comparison_45_vs_123 <- analysis_df |>
  filter(!is.na(strictness_group)) |>
  group_by(strictness_group) |>
  summarize(
    states = n(),
    mean_rate_per_100k = mean(rate_per_100k, na.rm = TRUE),
    median_rate_per_100k = median(rate_per_100k, na.rm = TRUE),
    mean_rate_per_capita = mean(rate_per_capita, na.rm = TRUE),
    median_rate_per_capita = median(rate_per_capita, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(strictness_group = factor(strictness_group, levels = c("Strict (4-5)", "Less strict (1-3)"))) |>
  arrange(strictness_group)

comparison_45_vs_123
```

## Heat maps (accessible palette)

## Gun-law strictness (Likert)

Accessibility color note:
This map uses a colorblind-safe, ordered categorical palette (A to F) so each law-grade band is distinguishable without relying on red/green contrast.

```{r map-laws}
p_laws <- plot_usmap(data = analysis_df, values = "law_band") +
  scale_fill_manual(
    values = c(
      "A" = "#0072B2",
      "B" = "#56B4E9",
      "C" = "#999999",
      "D" = "#E69F00",
      "F" = "#D55E00"
    ),
    limits = c("A", "B", "C", "D", "F"),
    breaks = c("A", "B", "C", "D", "F"),
    drop = FALSE,
    na.translate = FALSE,
    name = "Law grade band (A-F)"
  ) +
  labs(
    title = "Gun Law Strictness Map (A blue -> F red)",
    subtitle = "Colorblind-safe palette: A=blue, B=sky blue, C=gray, D=orange, F=orange-red"
  ) +
  theme_void() +
  theme(legend.position = "right")

p_laws
```

## Firearm mortality rate (CDC)

Accessibility color note:
This map uses a perceptually uniform `viridis` sequential scale so equal changes in mortality rate appear as equal visual steps for most viewers, including color-vision-deficient users.

```{r map-mortality}
p_mortality <- plot_usmap(data = analysis_df, values = "rate_per_100k") +
  scale_fill_viridis_c(option = "C", name = "Deaths per 100k") +
  labs(
    title = glue("CDC firearm mortality rate by state ({analysis_year})"),
    subtitle = glue("API: data.cdc.gov (SODA). Likert is from GIFFORDS grades (collapsed A–F → 1–5).")
  ) +
  theme_void() +
  theme(legend.position = "right")

p_mortality
```

## Combined map: strictness color + mortality distance scaling

Accessibility color note:
This map keeps the same colorblind-safe strictness palette and adds size encoding for mortality distance, so interpretation does not depend on color alone.

```{r map-combined-scaled}
map_group_mean_of_means <- analysis_df |>
  dplyr::filter(!is.na(strictness_group), !is.na(rate_per_100k)) |>
  dplyr::group_by(strictness_group) |>
  dplyr::summarize(group_mean = mean(rate_per_100k, na.rm = TRUE), .groups = "drop") |>
  dplyr::summarize(mean_of_means = mean(group_mean, na.rm = TRUE)) |>
  dplyr::pull(mean_of_means)

state_sf <- usmap::us_map(regions = "states")
state_coords <- as.data.frame(sf::st_coordinates(state_sf$geom))
state_attr <- state_sf |>
  sf::st_drop_geometry() |>
  dplyr::mutate(row_id = dplyr::row_number())

state_shapes <- state_coords |>
  dplyr::transmute(
    row_id = L3,
    group = paste(L3, L2, L1, sep = "-"),
    x = X,
    y = Y
  ) |>
  dplyr::left_join(
    state_attr |>
      dplyr::select(row_id, abbr),
    by = "row_id"
  ) |>
  dplyr::transmute(
    state_abbr = abbr,
    x = x,
    y = y,
    group = group
  ) |>
  dplyr::left_join(
    analysis_df |>
      dplyr::select(state_abbr, rate_per_100k, law_band),
    by = "state_abbr"
  ) |>
  dplyr::filter(state_abbr != "PR")

state_centroids <- state_shapes |>
  dplyr::group_by(state_abbr) |>
  dplyr::summarize(
    cx = mean(x, na.rm = TRUE),
    cy = mean(y, na.rm = TRUE),
    rate_per_100k = dplyr::first(rate_per_100k),
    law_band = dplyr::first(law_band),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    rel_from_mean = (rate_per_100k - map_group_mean_of_means) / map_group_mean_of_means,
    # Scale up/down from centroid by relative distance from mean of group means.
    scale_factor = pmin(pmax(1 + 0.7 * rel_from_mean, 0.55), 1.55)
  )

us_center <- state_centroids |>
  dplyr::summarize(ux = mean(cx, na.rm = TRUE), uy = mean(cy, na.rm = TRUE))

exploded_map <- state_shapes |>
  dplyr::left_join(state_centroids |>
                     dplyr::select(state_abbr, cx, cy, scale_factor),
                   by = "state_abbr") |>
  dplyr::mutate(
    # Explode outward from national center to reduce overlap.
    x_shift = 0.45 * (cx - us_center$ux),
    y_shift = 0.45 * (cy - us_center$uy),
    # Original-size (unscaled) exploded border coordinates.
    x_border = x + x_shift,
    y_border = y + y_shift,
    x2 = cx + (x - cx) * scale_factor + x_shift,
    y2 = cy + (y - cy) * scale_factor + y_shift
  )

p_combo <- ggplot() +
  geom_polygon(
    data = exploded_map,
    aes(x = x2, y = y2, group = group, fill = law_band),
    color = "white",
    linewidth = 0.2
  ) +
  geom_polygon(
    data = exploded_map,
    aes(x = x_border, y = y_border, group = group),
    fill = NA,
    color = "#00e676",
    linetype = "22",
    linewidth = 0.35
  ) +
  scale_fill_manual(
    values = c(
      "F" = "#D55E00",
      "D" = "#E69F00",
      "C" = "#999999",
      "B" = "#56B4E9",
      "A" = "#0072B2"
    ),
    limits = c("A", "B", "C", "D", "F"),
    breaks = c("A", "B", "C", "D", "F"),
    drop = FALSE,
    na.translate = FALSE,
    name = "Law grade band"
  ) +
  coord_equal() +
  labs(
    title = "Combined map: law strictness + mortality-distance resizing",
    subtitle = "State size changes by distance from the mean of group means; dashed outlines show original borders."
  ) +
  theme_void() +
  theme(legend.position = "right")

p_combo
```

## Year-by-year chart frames (2017 to latest)

The following sections generate the same 5 chart types for each available year.
Charts are grouped by chart type so you can animate each type over time.

```{r yearly-data-prep}
giffords_years <- giffords_grades |>
  dplyr::transmute(year = suppressWarnings(as.integer(year))) |>
  dplyr::filter(!is.na(year), year >= start_year) |>
  dplyr::pull(year) |>
  unique() |>
  sort()

cdc_firearm_cov <- cdc_state |>
  dplyr::transmute(period = suppressWarnings(as.integer(period)), state_abbr = to_state_abbr(name)) |>
  dplyr::filter(!is.na(period), !is.na(state_abbr), period >= start_year) |>
  dplyr::distinct(period, state_abbr) |>
  dplyr::count(period, name = "n_states")

cdc_proxy_cov <- cdc_state_raw |>
  dplyr::transmute(
    intent = as.character(intent),
    period = suppressWarnings(as.integer(period)),
    state_abbr = to_state_abbr(name)
  ) |>
  dplyr::filter(intent == "All_Homicide", !is.na(period), !is.na(state_abbr), period >= start_year) |>
  dplyr::distinct(period, state_abbr) |>
  dplyr::count(period, name = "n_states")

panel_years <- sort(intersect(
  giffords_years,
  sort(unique(c(
    cdc_firearm_cov |> dplyr::filter(n_states >= 45) |> dplyr::pull(period),
    cdc_proxy_cov |> dplyr::filter(n_states >= 45) |> dplyr::pull(period)
  )))
))

if (length(panel_years) == 0) stop("No overlapping years with >=45 states for yearly charts.")

build_year_df <- function(target_year) {
  # Primary series: same FA_Deaths logic as main analysis (preserves recent-year behavior).
  state_year <- cdc_state |>
    dplyr::filter(period == target_year) |>
    dplyr::transmute(
      state_abbr = to_state_abbr(name),
      rate_per_100k = suppressWarnings(as.numeric(rate))
    ) |>
    dplyr::filter(!is.na(state_abbr), !is.na(rate_per_100k)) |>
    dplyr::group_by(state_abbr) |>
    dplyr::summarize(rate_per_100k = mean(rate_per_100k, na.rm = TRUE), .groups = "drop")
  metric_basis <- "FA_Deaths"

  # Fallback for early years where firearm series is unavailable in current endpoint.
  if (nrow(state_year) < 45) {
    state_year <- cdc_state_raw |>
      dplyr::transmute(
        intent = as.character(intent),
        period = suppressWarnings(as.integer(period)),
        state_abbr = to_state_abbr(name),
        rate_per_100k = suppressWarnings(as.numeric(rate))
      ) |>
      dplyr::filter(
        intent == "All_Homicide",
        period == target_year,
        !is.na(state_abbr),
        !is.na(rate_per_100k)
      ) |>
      dplyr::group_by(state_abbr) |>
      dplyr::summarize(rate_per_100k = mean(rate_per_100k, na.rm = TRUE), .groups = "drop")
    metric_basis <- "Proxy: All_Homicide"
  }

  law_year <- giffords_grades |>
    dplyr::filter(year == target_year) |>
    dplyr::transmute(
      state_abbr = to_state_abbr(state),
      grade = as.character(grade),
      likert = as.numeric(likert)
    ) |>
    dplyr::filter(!is.na(state_abbr), !is.na(likert))

  out <- state_year |>
    dplyr::left_join(law_year, by = "state_abbr") |>
    dplyr::filter(!is.na(rate_per_100k), !is.na(likert), state_abbr != "PR") |>
    dplyr::mutate(
      state = state_abbr,
      metric_basis = metric_basis,
      law_band = factor(likert, levels = 1:5, labels = c("F", "D", "C", "B", "A"), ordered = TRUE),
      strictness_group = dplyr::case_when(
        likert %in% c(4, 5) ~ "Strict (4-5)",
        likert %in% c(1, 2, 3) ~ "Less strict (1-3)",
        TRUE ~ NA_character_
      )
    )

  if (nrow(out) < 10) return(NULL)
  out
}

yearly_analysis <- setNames(vector("list", length(panel_years)), as.character(panel_years))
for (yr in panel_years) yearly_analysis[[as.character(yr)]] <- build_year_df(yr)
yearly_plot_laws <- list()
yearly_plot_mortality <- list()
yearly_plot_combined <- list()
yearly_plot_weighted <- list()
yearly_plot_violin <- list()

cat("Yearly chart frame years:\n")
print(panel_years)

yearly_quality <- tibble::tibble(year = panel_years) |>
  dplyr::mutate(
    states_after_join = purrr::map_int(year, ~{
      df <- yearly_analysis[[as.character(.x)]]
      if (is.null(df)) 0L else nrow(df)
    }),
    metric_basis = purrr::map_chr(year, ~{
      df <- yearly_analysis[[as.character(.x)]]
      if (is.null(df) || !("metric_basis" %in% names(df))) NA_character_ else unique(df$metric_basis)[1]
    })
  )
print(yearly_quality, n = Inf)

panel_rate_max <- max(
  purrr::map_dbl(yearly_analysis, ~ if (is.null(.x)) NA_real_ else max(.x$rate_per_100k, na.rm = TRUE)),
  na.rm = TRUE
)
panel_rate_max <- ceiling(panel_rate_max / 10) * 10

state_sf_panel <- usmap::us_map(regions = "states")
state_coords_panel <- as.data.frame(sf::st_coordinates(state_sf_panel$geom))
state_attr_panel <- state_sf_panel |>
  sf::st_drop_geometry() |>
  dplyr::mutate(row_id = dplyr::row_number())

base_state_shapes <- state_coords_panel |>
  dplyr::transmute(
    row_id = L3,
    group = paste(L3, L2, L1, sep = "-"),
    x = X,
    y = Y
  ) |>
  dplyr::left_join(
    state_attr_panel |>
      dplyr::select(row_id, abbr),
    by = "row_id"
  ) |>
  dplyr::transmute(state_abbr = abbr, x = x, y = y, group = group) |>
  dplyr::filter(state_abbr != "PR")
```

### 1) Gun-law strictness map by year

Accessibility color note:
We use a colorblind-safe categorical palette with strong hue separation (blue/sky/gray/orange/orange-red) so adjacent grade bands remain distinguishable for common color-vision deficiencies.

```{r yearly-laws-map, fig.height=5.5, fig.width=10}
for (yr in panel_years) {
  df <- yearly_analysis[[as.character(yr)]]
  if (is.null(df)) next
  basis_txt <- paste(unique(df$metric_basis), collapse = ", ")

  p <- plot_usmap(data = df, values = "law_band") +
    scale_fill_manual(
      values = c("F" = "#D55E00", "D" = "#E69F00", "C" = "#999999", "B" = "#56B4E9", "A" = "#0072B2"),
      limits = c("A", "B", "C", "D", "F"),
      breaks = c("A", "B", "C", "D", "F"),
      drop = FALSE,
      na.translate = FALSE,
      name = "Law grade band (A-F)"
    ) +
    labs(title = glue("Gun Law Strictness Map (A blue -> F red) ({yr})"),
         subtitle = glue("Colorblind-safe palette (A→F) • Mortality source: {basis_txt}")) +
    theme_void() +
    theme(legend.position = "right")

  print(p)
  yearly_plot_laws[[as.character(yr)]] <- p
}
```

### 2) Mortality map by year

Accessibility color note:
We use `viridis` (option C), a perceptually uniform, colorblind-safe sequential scale so equal numeric changes are visually consistent and legible.

```{r yearly-mortality-map, fig.height=5.5, fig.width=10}
for (yr in panel_years) {
  df <- yearly_analysis[[as.character(yr)]]
  if (is.null(df)) next
  basis_txt <- paste(unique(df$metric_basis), collapse = ", ")

  p <- plot_usmap(data = df, values = "rate_per_100k") +
    scale_fill_viridis_c(option = "C", name = "Deaths per 100k", limits = c(0, panel_rate_max)) +
    labs(
      title = glue("Mortality rate by state ({yr})"),
      subtitle = glue("Source: {basis_txt}")
    ) +
    theme_void() +
    theme(legend.position = "right")

  print(p)
  yearly_plot_mortality[[as.character(yr)]] <- p
}
```

### 3) Combined map by year (color + resized polygons)

Accessibility color note:
Strictness uses the same colorblind-safe categorical palette for consistency, and the second encoding (state size) does not rely on color perception alone.

```{r yearly-combined-map, fig.height=5.5, fig.width=10}
for (yr in panel_years) {
  df <- yearly_analysis[[as.character(yr)]]
  if (is.null(df)) next
  basis_txt <- paste(unique(df$metric_basis), collapse = ", ")

  group_mean_of_means_year <- df |>
    dplyr::group_by(strictness_group) |>
    dplyr::summarize(group_mean = mean(rate_per_100k, na.rm = TRUE), .groups = "drop") |>
    dplyr::summarize(mom = mean(group_mean, na.rm = TRUE)) |>
    dplyr::pull(mom)

  shapes <- base_state_shapes |>
    dplyr::left_join(df |> dplyr::select(state_abbr, rate_per_100k, law_band), by = "state_abbr")

  cent <- shapes |>
    dplyr::group_by(state_abbr) |>
    dplyr::summarize(
      cx = mean(x, na.rm = TRUE),
      cy = mean(y, na.rm = TRUE),
      rate_per_100k = dplyr::first(rate_per_100k),
      law_band = dplyr::first(law_band),
      .groups = "drop"
    ) |>
    dplyr::mutate(
      rel_from_mean = (rate_per_100k - group_mean_of_means_year) / group_mean_of_means_year,
      scale_factor = pmin(pmax(1 + 0.7 * rel_from_mean, 0.55), 1.55)
    )

  center_us <- cent |> dplyr::summarize(ux = mean(cx, na.rm = TRUE), uy = mean(cy, na.rm = TRUE))

  exploded <- shapes |>
    dplyr::left_join(cent |> dplyr::select(state_abbr, cx, cy, scale_factor), by = "state_abbr") |>
    dplyr::mutate(
      x_shift = 0.45 * (cx - center_us$ux),
      y_shift = 0.45 * (cy - center_us$uy),
      x_border = x + x_shift,
      y_border = y + y_shift,
      x2 = cx + (x - cx) * scale_factor + x_shift,
      y2 = cy + (y - cy) * scale_factor + y_shift
    )

  p <- ggplot() +
    geom_polygon(data = exploded, aes(x = x2, y = y2, group = group, fill = law_band), color = "white", linewidth = 0.2) +
    geom_polygon(data = exploded, aes(x = x_border, y = y_border, group = group), fill = NA, color = "#00e676", linetype = "22", linewidth = 0.35) +
    scale_fill_manual(
      values = c("F" = "#D55E00", "D" = "#E69F00", "C" = "#999999", "B" = "#56B4E9", "A" = "#0072B2"),
      limits = c("A", "B", "C", "D", "F"),
      breaks = c("A", "B", "C", "D", "F"),
      drop = FALSE, na.translate = FALSE, name = "Law grade band (A-F)"
    ) +
    coord_equal() +
    labs(title = glue("Combined map ({yr})"),
         subtitle = glue("Color = strictness; size = distance from mean of group means • Source: {basis_txt}")) +
    theme_void() +
    theme(legend.position = "right")

  print(p)
  yearly_plot_combined[[as.character(yr)]] <- p
}
```

### 4) Weighted decomposition by year

Accessibility color note:
Group fills use high-contrast, colorblind-safe hues (blue vs orange-red), and the mean reference is additionally encoded by dashed line/label, not color alone.

```{r yearly-weighted-decomp, fig.height=5.5, fig.width=10}
for (yr in panel_years) {
  df <- yearly_analysis[[as.character(yr)]]
  if (is.null(df)) next
  basis_txt <- paste(unique(df$metric_basis), collapse = ", ")

  gs <- df |>
    dplyr::group_by(strictness_group) |>
    dplyr::summarize(states = n(), mean_rate_per_100k = mean(rate_per_100k, na.rm = TRUE), .groups = "drop") |>
    dplyr::mutate(strictness_group = factor(strictness_group, levels = c("Strict (4-5)", "Less strict (1-3)"))) |>
    dplyr::arrange(strictness_group) |>
    dplyr::mutate(xmin = dplyr::lag(cumsum(states), default = 0), xmax = cumsum(states))

  mom <- mean(gs$mean_rate_per_100k, na.rm = TRUE)

  p <- ggplot(gs) +
    geom_rect(aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = mean_rate_per_100k, fill = strictness_group), color = "white") +
    geom_hline(yintercept = mom, linetype = "dashed", linewidth = 1, color = "black") +
    annotate("text", x = sum(gs$states) + 1, y = mom + 0.35, label = paste0(round(mom, 1), " mean"), hjust = 0, vjust = 0, size = 3.1) +
    scale_fill_manual(values = c("Less strict (1-3)" = "#D55E00", "Strict (4-5)" = "#0072B2")) +
    scale_x_continuous(limits = c(0, sum(gs$states) + 3), breaks = seq(0, sum(gs$states), by = 10), minor_breaks = NULL) +
    labs(title = glue("Weighted decomposition ({yr})"),
         subtitle = glue("Dashed line/label = mean of group means • Source: {basis_txt}"),
         x = "Cumulative state count", y = "Group mean deaths per 100k", fill = "Group") +
    theme_minimal() +
    theme(panel.grid = element_blank())

  print(p)
  yearly_plot_weighted[[as.character(yr)]] <- p
}
```

### 5) Violin + quantile box by year

Accessibility color note:
Group fills use colorblind-safe contrast, while medians, quantile boxes, and point markers provide shape/position cues independent of color.

```{r yearly-violin, fig.height=5.5, fig.width=10, results='asis'}
for (yr in panel_years) {
  df <- yearly_analysis[[as.character(yr)]]
  if (is.null(df)) next
  basis_txt <- paste(unique(df$metric_basis), collapse = ", ")

  gs_mom <- df |>
    dplyr::group_by(strictness_group) |>
    dplyr::summarize(group_mean = mean(rate_per_100k, na.rm = TRUE), .groups = "drop") |>
    dplyr::summarize(mom = mean(group_mean, na.rm = TRUE)) |>
    dplyr::pull(mom)

  low_df <- df |> dplyr::group_by(strictness_group) |> dplyr::slice_min(order_by = rate_per_100k, n = 1, with_ties = FALSE) |> dplyr::ungroup()
  med_df <- df |>
    dplyr::group_by(strictness_group) |>
    dplyr::mutate(grp_median = median(rate_per_100k, na.rm = TRUE), med_dist = abs(rate_per_100k - grp_median)) |>
    dplyr::slice_min(order_by = med_dist, n = 1, with_ties = FALSE) |>
    dplyr::ungroup() |>
    dplyr::select(-grp_median, -med_dist)
  high_df <- df |> dplyr::group_by(strictness_group) |> dplyr::slice_max(order_by = rate_per_100k, n = 1, with_ties = FALSE) |> dplyr::ungroup()
  v_labels <- dplyr::bind_rows(low_df, med_df, high_df) |> dplyr::mutate(state_abbr = toupper(state_abbr))

  p <- ggplot(df, aes(x = strictness_group, y = rate_per_100k, fill = strictness_group)) +
    geom_violin(alpha = 0.35, trim = FALSE) +
    geom_boxplot(width = 0.2, outlier.shape = NA) +
    geom_point(data = v_labels, color = "gray45", size = 2.5) +
    geom_text(data = v_labels, aes(label = state_abbr), color = "gray30", nudge_x = 0.25, hjust = 0, vjust = 0, size = 3.5, show.legend = FALSE) +
    stat_summary(fun = mean, geom = "point", shape = 21, size = 3, fill = "yellow") +
    geom_hline(yintercept = gs_mom, linetype = "dashed", linewidth = 1, color = "black") +
    labs(
      title = glue("Distribution by group with quantile box ({yr})"),
      subtitle = glue("Source: {basis_txt}"),
      x = "Strictness group",
      y = "Deaths per 100k",
      fill = "Group"
    ) +
    scale_y_continuous(breaks = seq(0, panel_rate_max, by = 10), minor_breaks = NULL) +
    theme_minimal() +
    theme(
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_line(color = "gray80", linewidth = 0.3)
    )

  print(p)
  yearly_plot_violin[[as.character(yr)]] <- p

  dist_tbl <- df |>
    dplyr::filter(!is.na(strictness_group), !is.na(rate_per_100k)) |>
    dplyr::group_by(strictness_group) |>
    dplyr::summarize(
      n = dplyr::n(),
      mean_rate = mean(rate_per_100k, na.rm = TRUE),
      median_rate = median(rate_per_100k, na.rm = TRUE),
      skew = {
        m <- mean(rate_per_100k, na.rm = TRUE)
        s <- sd(rate_per_100k, na.rm = TRUE)
        if (is.na(s) || s == 0) NA_real_ else mean((rate_per_100k - m)^3, na.rm = TRUE) / (s^3)
      },
      shapiro_p = if (dplyr::n() >= 3 && dplyr::n() <= 5000) shapiro.test(rate_per_100k)$p.value else NA_real_,
      .groups = "drop"
    ) |>
    dplyr::mutate(
      distribution_class = dplyr::case_when(
        is.na(skew) ~ "Undetermined",
        abs(skew) < 0.2 & !is.na(shapiro_p) & shapiro_p > 0.05 ~ "Approximately normal",
        skew >= 0.75 ~ "Right skew",
        skew >= 0.2 ~ "Mild right skew",
        skew <= -0.75 ~ "Left skew",
        skew <= -0.2 ~ "Mild left skew",
        TRUE ~ "Non-normal, near symmetric"
      ),
      year = yr
    ) |>
    dplyr::select(strictness_group, distribution_class, year, n, mean_rate, median_rate, skew, shapiro_p) |>
    dplyr::mutate(
      strictness_group = factor(strictness_group, levels = c("Strict (4-5)", "Less strict (1-3)"))
    ) |>
    dplyr::arrange(strictness_group) |>
    dplyr::mutate(
      mean_rate = round(mean_rate, 2),
      median_rate = round(median_rate, 2),
      skew = round(skew, 3),
      shapiro_p = round(shapiro_p, 4)
    )

  print(
    knitr::kable(
      dist_tbl,
      caption = glue("Distribution classification ({yr})"),
      align = c("l", "l", "c", "r", "r", "r", "r", "r")
    )
  )
  cat("\n\n")
}
```

## Visualizing group means (mean-of-means reference)

Accessibility color note:
These group comparison charts use a high-contrast blue vs orange-red pair and also encode meaning with line style, point shape, and position so key comparisons remain readable without color cues.

```{r mean-diff-visuals, fig.height=5.5, fig.width=10}
group_summary <- analysis_df |>
  filter(!is.na(strictness_group)) |>
  group_by(strictness_group) |>
  summarize(
    states = n(),
    mean_rate_per_100k = mean(rate_per_100k, na.rm = TRUE),
    median_rate_per_100k = median(rate_per_100k, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(strictness_group = factor(strictness_group, levels = c("Strict (4-5)", "Less strict (1-3)")))

overall_weighted_mean <- mean(analysis_df$rate_per_100k, na.rm = TRUE)
group_mean_of_means <- mean(group_summary$mean_rate_per_100k, na.rm = TRUE)

mean_compare <- tibble::tibble(
  metric = "Deaths per 100k",
  mean_of_group_means = group_mean_of_means,
  weighted = overall_weighted_mean
)

mean_compare

# 2) Group-size weighted bar decomposition
decomp <- group_summary |>
  arrange(strictness_group) |>
  mutate(
    xmin = dplyr::lag(cumsum(states), default = 0),
    xmax = cumsum(states)
  )

p_weighted_decomp <- ggplot(decomp) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = mean_rate_per_100k, fill = strictness_group), color = "white") +
  geom_hline(yintercept = group_mean_of_means, linetype = "dashed", linewidth = 1, color = "black") +
  annotate(
    "text",
    x = sum(decomp$states) + 1,
    y = group_mean_of_means + 0.35,
    label = paste0(round(group_mean_of_means, 1), " mean"),
    hjust = 0,
    vjust = 0,
    size = 3.1
  ) +
  scale_fill_manual(
    values = c(
      "Less strict (1-3)" = "#D55E00",
      "Strict (4-5)" = "#0072B2"
    )
  ) +
  scale_x_continuous(
    limits = c(0, sum(decomp$states) + 3),
    breaks = seq(0, sum(decomp$states), by = 10),
    minor_breaks = NULL
  ) +
  labs(
    title = "2) Weighted decomposition (bar width = group share)",
    subtitle = "Dashed line and label indicate the mean of group means.",
    x = "Cumulative state count",
    y = "Group mean deaths per 100k",
    fill = "Group"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())

p_weighted_decomp

# 6) Violin + box by group with mean markers
lowest_df <- analysis_df |>
  filter(!is.na(strictness_group), !is.na(rate_per_100k), !is.na(state_abbr)) |>
  group_by(strictness_group) |>
  slice_min(order_by = rate_per_100k, n = 1, with_ties = FALSE) |>
  ungroup()

median_df <- analysis_df |>
  filter(!is.na(strictness_group), !is.na(rate_per_100k), !is.na(state_abbr)) |>
  group_by(strictness_group) |>
  mutate(
    grp_median = median(rate_per_100k, na.rm = TRUE),
    med_dist = abs(rate_per_100k - grp_median)
  ) |>
  slice_min(order_by = med_dist, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(-grp_median, -med_dist)

highest_df <- analysis_df |>
  filter(!is.na(strictness_group), !is.na(rate_per_100k), !is.na(state_abbr)) |>
  group_by(strictness_group) |>
  slice_max(order_by = rate_per_100k, n = 1, with_ties = FALSE) |>
  ungroup()

violin_labels <- bind_rows(
  lowest_df |> mutate(label_type = "low"),
  median_df |> mutate(label_type = "median"),
  highest_df |> mutate(label_type = "high")
) |>
  group_by(strictness_group, label_type) |>
  slice_head(n = 1) |>
  ungroup() |>
  mutate(state_abbr = toupper(state_abbr))

p_violin <- ggplot(analysis_df, aes(x = strictness_group, y = rate_per_100k, fill = strictness_group)) +
  geom_violin(alpha = 0.35, trim = FALSE) +
  geom_boxplot(width = 0.2, outlier.shape = NA) +
  geom_point(data = violin_labels, color = "gray45", size = 2.5) +
  geom_text(
    data = violin_labels,
    aes(label = state_abbr),
    color = "gray30",
    nudge_x = 0.25,
    hjust = 0,
    vjust = 0,
    size = 3.5,
    show.legend = FALSE
  ) +
  stat_summary(fun = mean, geom = "point", shape = 21, size = 3, fill = "yellow") +
  geom_hline(yintercept = group_mean_of_means, linetype = "dashed", linewidth = 1, color = "black") +
  labs(
    title = "6) Distribution by group with quantile box (IQR)",
    x = "Strictness group",
    y = "Deaths per 100k",
    fill = "Group"
  ) +
  scale_y_continuous(
    breaks = seq(0, ceiling(max(analysis_df$rate_per_100k, na.rm = TRUE) / 10) * 10, by = 10),
    minor_breaks = NULL
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(color = "gray80", linewidth = 0.3)
  )

p_violin

distribution_classification <- analysis_df |>
  filter(!is.na(strictness_group), !is.na(rate_per_100k)) |>
  group_by(strictness_group) |>
  summarize(
    n = n(),
    mean_rate = mean(rate_per_100k, na.rm = TRUE),
    median_rate = median(rate_per_100k, na.rm = TRUE),
    sd_rate = sd(rate_per_100k, na.rm = TRUE),
    skew = {
      m <- mean(rate_per_100k, na.rm = TRUE)
      s <- sd(rate_per_100k, na.rm = TRUE)
      if (is.na(s) || s == 0) NA_real_ else mean((rate_per_100k - m)^3, na.rm = TRUE) / (s^3)
    },
    shapiro_p = if (n() >= 3 && n() <= 5000) shapiro.test(rate_per_100k)$p.value else NA_real_,
    .groups = "drop"
  ) |>
  mutate(
    distribution_class = dplyr::case_when(
      is.na(skew) ~ "Undetermined",
      abs(skew) < 0.2 & !is.na(shapiro_p) & shapiro_p > 0.05 ~ "Approximately normal",
      skew >= 0.75 ~ "Right skew",
      skew >= 0.2 ~ "Mild right skew",
      skew <= -0.75 ~ "Left skew",
      skew <= -0.2 ~ "Mild left skew",
      TRUE ~ "Non-normal, near symmetric"
    )
  ) |>
  select(strictness_group, distribution_class, n, mean_rate, median_rate, skew, shapiro_p)

distribution_classification
```

## Correlation and regression (association, not causation)

```{r stats}
# cor.test needs at least 2 finite observations; we already sanity-check above,
# but keep a guard so knitting fails gracefully if inputs change.
if (sum(is.finite(analysis_df$likert) & is.finite(analysis_df$rate_per_100k)) < 3) {
  stop("Not enough finite observations for correlation/regression after filtering.")
}

cor_spear <- cor.test(analysis_df$likert, analysis_df$rate_per_100k, method = "spearman")
lm_fit <- lm(rate_per_100k ~ likert, data = analysis_df)

cor_spear
summary(lm_fit)
```

# Export all charts to files (PNG + SVG)

```{r export-charts}
export_dir <- "imaages"
dir.create(export_dir, showWarnings = FALSE, recursive = TRUE)

p_scatter_export <- ggplot(analysis_df, aes(x = likert, y = rate_per_100k)) +
  geom_jitter(width = 0.2, height = 0, size = 3, alpha = 0.85) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(breaks = 1:5) +
  labs(
    title = "Association: strictness vs mortality",
    subtitle = glue("Spearman rho = {round(unname(cor_spear$estimate), 2)} (p={signif(cor_spear$p.value, 3)})"),
    x = "Strictness (1 = lax … 5 = strict)",
    y = "Deaths per 100k"
  ) +
  theme_minimal()

main_plots <- list(
  "main_01_laws_heatmap" = p_laws,
  "main_02_mortality_heatmap" = p_mortality,
  "main_03_combined_heatmap" = p_combo,
  "main_04_scatter_association" = p_scatter_export,
  "main_05_weighted_decomposition" = p_weighted_decomp,
  "main_06_violin_quantile_box" = p_violin
)

save_plot_pair <- function(plot_obj, stem, width = 11, height = 6) {
  ggplot2::ggsave(
    filename = file.path(export_dir, paste0(stem, ".png")),
    plot = plot_obj,
    width = width,
    height = height,
    dpi = 320
  )
  ggplot2::ggsave(
    filename = file.path(export_dir, paste0(stem, ".svg")),
    plot = plot_obj,
    width = width,
    height = height,
    device = "svg"
  )
}

for (nm in names(main_plots)) {
  save_plot_pair(main_plots[[nm]], nm)
}

for (yr in panel_years) {
  y <- as.character(yr)
  if (!is.null(yearly_plot_laws[[y]])) {
    save_plot_pair(yearly_plot_laws[[y]], paste0("year_", y, "_01_laws_heatmap"))
  }
  if (!is.null(yearly_plot_mortality[[y]])) {
    save_plot_pair(yearly_plot_mortality[[y]], paste0("year_", y, "_02_mortality_heatmap"))
  }
  if (!is.null(yearly_plot_combined[[y]])) {
    save_plot_pair(yearly_plot_combined[[y]], paste0("year_", y, "_03_combined_heatmap"))
  }
  if (!is.null(yearly_plot_weighted[[y]])) {
    save_plot_pair(yearly_plot_weighted[[y]], paste0("year_", y, "_04_weighted_decomposition"))
  }
  if (!is.null(yearly_plot_violin[[y]])) {
    save_plot_pair(yearly_plot_violin[[y]], paste0("year_", y, "_05_violin_quantile_box"))
  }
}

list.files(export_dir) |> sort() |> head(20)
```

# PPT generation (Path B): officer + rvg

This section creates a PowerPoint file with **editable vector charts**.
After creating the PPT, open it in PowerPoint and **Ungroup** the chart objects (often twice) to animate components.

```{r make-ppt}
ppt <- read_pptx()

# Slide 1: Title / research question
ppt <- ppt |>
  add_slide(layout = "Title Slide", master = "Office Theme") |>
  ph_with(
    "Do Stricter Gun Laws Reduce Firearm Mortality?",
    location = ph_location_type(type = "ctrTitle")
  ) |>
  ph_with(
    glue("CDC API + GIFFORDS Likert • Latest year: {analysis_year}"),
    location = ph_location_type(type = "subTitle")
  )

# Slide 2: Assignment framing
assignment_text <- paste(
  "Story goal:",
  "Assess whether stricter state gun laws are associated with lower firearm mortality.",
  "",
  "Assignment requirements covered:",
  "1) CDC firearm mortality data via API (data.cdc.gov / SODA)",
  "2) 5-point Likert scale for gun-law strictness",
  "3) Determine whether stricter laws are linked to reduced deaths",
  "4) Present findings with heat maps using accessible color palettes",
  sep = "\n"
)

ppt <- ppt |>
  add_slide(layout = "Title and Content", master = "Office Theme") |>
  ph_with("Assignment Context & Question", location = ph_location_type(type = "title")) |>
  ph_with(assignment_text, location = ph_location_type(type = "body"))

# Slide 3: Data + Likert method
method_text <- paste(
  "Data sources:",
  "• CDC state firearm mortality rates per 100,000 (SODA API endpoint: fpsi-y8tj)",
  "• GIFFORDS annual state gun-law grades (A+ to F), scraped and cached",
  "",
  "5-point Likert mapping (strictness):",
  "• 5 = A+, A, A-",
  "• 4 = B+, B, B-",
  "• 3 = C+, C, C-",
  "• 2 = D+, D, D-",
  "• 1 = F",
  "",
  glue("Analysis year used: {analysis_year} (latest overlap with full state coverage)"),
  sep = "\n"
)

ppt <- ppt |>
  add_slide(layout = "Title and Content", master = "Office Theme") |>
  ph_with("Data Pipeline & Likert Construction", location = ph_location_type(type = "title")) |>
  ph_with(method_text, location = ph_location_type(type = "body"))

# Slide 4: Laws heatmap (accessible palette)
ppt <- ppt |>
  add_slide(layout = "Title and Content", master = "Office Theme") |>
  ph_with("Heat Map 1: Gun-Law Strictness (Likert 1–5)", location = ph_location_type(type = "title")) |>
  ph_with(dml(ggobj = p_laws), location = ph_location_type(type = "body"))

# Slide 5: Mortality heatmap (accessible palette)
ppt <- ppt |>
  add_slide(layout = "Title and Content", master = "Office Theme") |>
  ph_with("Heat Map 2: Firearm Mortality by State (CDC)", location = ph_location_type(type = "title")) |>
  ph_with(dml(ggobj = p_mortality), location = ph_location_type(type = "body"))

# Slide 6: Combined map
ppt <- ppt |>
  add_slide(layout = "Title and Content", master = "Office Theme") |>
  ph_with("Heat Map 3: Combined Strictness + Mortality Distance", location = ph_location_type(type = "title")) |>
  ph_with(dml(ggobj = p_combo), location = ph_location_type(type = "body"))

# Slide 7: Association plot
p_scatter <- ggplot(analysis_df, aes(x = likert, y = rate_per_100k)) +
  geom_jitter(width = 0.2, height = 0, size = 3, alpha = 0.85) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(breaks = 1:5) +
  labs(
    title = "Association: strictness vs mortality",
    subtitle = glue("Spearman rho = {round(unname(cor_spear$estimate), 2)} (p={signif(cor_spear$p.value, 3)})"),
    x = "Strictness (1 = lax … 5 = strict)",
    y = "Deaths per 100k"
  ) +
  theme_minimal()

ppt <- ppt |>
  add_slide(layout = "Title and Content", master = "Office Theme") |>
  ph_with("Association Check: Strictness vs Mortality", location = ph_location_type(type = "title")) |>
  ph_with(dml(ggobj = p_scatter), location = ph_location_type(type = "body"))

# Slides 8+: Export all yearly charts to PPT
if (exists("panel_years") && length(panel_years) > 0) {
  for (yr in panel_years) {
    y <- as.character(yr)

    if (!is.null(yearly_plot_laws[[y]])) {
      ppt <- ppt |>
        add_slide(layout = "Title and Content", master = "Office Theme") |>
        ph_with(glue("Year {yr}: Gun-Law Strictness Heat Map"), location = ph_location_type(type = "title")) |>
        ph_with(dml(ggobj = yearly_plot_laws[[y]]), location = ph_location_type(type = "body"))
    }

    if (!is.null(yearly_plot_mortality[[y]])) {
      ppt <- ppt |>
        add_slide(layout = "Title and Content", master = "Office Theme") |>
        ph_with(glue("Year {yr}: Firearm Mortality Heat Map"), location = ph_location_type(type = "title")) |>
        ph_with(dml(ggobj = yearly_plot_mortality[[y]]), location = ph_location_type(type = "body"))
    }

    if (!is.null(yearly_plot_combined[[y]])) {
      ppt <- ppt |>
        add_slide(layout = "Title and Content", master = "Office Theme") |>
        ph_with(glue("Year {yr}: Combined Strictness + Mortality Map"), location = ph_location_type(type = "title")) |>
        ph_with(dml(ggobj = yearly_plot_combined[[y]]), location = ph_location_type(type = "body"))
    }

    if (!is.null(yearly_plot_weighted[[y]])) {
      ppt <- ppt |>
        add_slide(layout = "Title and Content", master = "Office Theme") |>
        ph_with(glue("Year {yr}: Weighted Decomposition"), location = ph_location_type(type = "title")) |>
        ph_with(dml(ggobj = yearly_plot_weighted[[y]]), location = ph_location_type(type = "body"))
    }

    if (!is.null(yearly_plot_violin[[y]])) {
      ppt <- ppt |>
        add_slide(layout = "Title and Content", master = "Office Theme") |>
        ph_with(glue("Year {yr}: Violin + Quantile Box"), location = ph_location_type(type = "title")) |>
        ph_with(dml(ggobj = yearly_plot_violin[[y]]), location = ph_location_type(type = "body"))
    }
  }
}

# Final slide: Results + conclusion
result_direction <- ifelse(unname(cor_spear$estimate) < 0, "negative", "positive")

conclusion_text <- paste(
  "Key result:",
  glue("• Spearman rho = {round(unname(cor_spear$estimate), 2)} ({result_direction} association), p = {signif(cor_spear$p.value, 3)}"),
  "",
  "Interpretation:",
  "• In this cross-sectional state-year view, higher strictness tends to align with lower mortality when rho is negative.",
  "• This supports an association, not causal proof.",
  "",
  "Limitations:",
  "• Potential confounding (demographics, urbanicity, poverty, enforcement, etc.)",
  "• Policy endogeneity and timing effects",
  "• Single-year comparison for the core assignment question",
  sep = "\n"
)

ppt <- ppt |>
  add_slide(layout = "Title and Content", master = "Office Theme") |>
  ph_with("Conclusion: Do Stricter Laws Reduce Mortality?", location = ph_location_type(type = "title")) |>
  ph_with(conclusion_text, location = ph_location_type(type = "body"))

out_pptx <- "Gun_Laws_CDC_GIFFORDS.pptx"
print(ppt, target = out_pptx)

out_pptx
```

# How to animate in PowerPoint

1. Open the generated `.pptx`
2. Click a chart → **Right click → Group → Ungroup** (repeat if needed)
3. Now you can animate:
   - individual map layers / legends / text
   - scatter points separate from trend line
   - chart elements and labels

# Next enhancements (optional)

- Add a **panel** across years (1999–present) with a mixed model or state fixed effects
- Add outlier slides highlighting states with strict laws but high rates (and vice versa)

# Optional: probe how far back GIFFORDS is parseable

Run this chunk manually if you want to test earliest available scorecard years.

```{r giffords-year-probe, eval=FALSE}
scorecard_url <- function(year) {
  if (year == 2024) {
    "https://giffords.org/resources/scorecard2024/"
  } else {
    paste0("https://giffords.org/lawcenter/resources/scorecard", year, "/")
  }
}

grade_to_likert <- function(g) {
  g <- stringr::str_trim(g)
  dplyr::case_when(
    stringr::str_detect(g, "^A") ~ 5,
    stringr::str_detect(g, "^B") ~ 4,
    stringr::str_detect(g, "^C") ~ 3,
    stringr::str_detect(g, "^D") ~ 2,
    stringr::str_detect(g, "^F") ~ 1,
    TRUE ~ NA_real_
  )
}

scrape_scorecard_year <- function(year) {
  url <- scorecard_url(year)
  page <- rvest::read_html(url)

  tabs <- page |> rvest::html_elements("table") |> rvest::html_table(fill = TRUE)
  if (length(tabs) > 0) {
    cand <- tabs[[which.max(purrr::map_int(tabs, nrow))]] |> janitor::clean_names()

    if (any(stringr::str_detect(names(cand), "grade")) &&
        any(stringr::str_detect(names(cand), "state|name"))) {

      state_col <- names(cand)[which(stringr::str_detect(names(cand), "state|name"))[1]]
      grade_col <- names(cand)[which(stringr::str_detect(names(cand), "grade"))[1]]

      out <- cand |>
        dplyr::transmute(
          state = as.character(.data[[state_col]]) |> stringr::str_squish(),
          grade = as.character(.data[[grade_col]]) |> stringr::str_squish(),
          year  = year,
          likert = grade_to_likert(grade)
        ) |>
        dplyr::filter(state %in% state.name) |>
        dplyr::distinct()

      if (nrow(out) >= 45) return(out)
    }
  }

  txt <- page |> rvest::html_text2()
  pos <- stringr::str_locate(txt, stringr::fixed("Gun Law Strength", ignore_case = TRUE))[1,1]
  if (is.na(pos)) stop("marker_not_found")

  tail_txt <- stringr::str_sub(txt, pos, pos + 30000)
  tail_txt <- stringr::str_replace_all(tail_txt, "【[^】]*】", " ")
  tail_txt <- stringr::str_squish(tail_txt)

  m <- stringr::str_match_all(
    tail_txt,
    "\\b\\d{1,2}\\s+([A-Za-z ]+?)\\s+(A\\+|A-|A|B\\+|B-|B|C\\+|C-|C|D\\+|D-|D|F)\\b"
  )[[1]]

  if (nrow(m) == 0) stop("no_pairs")

  tibble::tibble(
    state = stringr::str_squish(m[,2]),
    grade = m[,3],
    year = year
  ) |>
    dplyr::filter(state %in% state.name) |>
    dplyr::mutate(likert = grade_to_likert(grade)) |>
    dplyr::distinct()
}

years_to_test <- 2008:2024

probe <- purrr::map_dfr(years_to_test, function(y) {
  u <- scorecard_url(y)
  out <- tryCatch(scrape_scorecard_year(y), error = function(e) e)
  if (inherits(out, "error")) {
    tibble::tibble(year = y, ok = FALSE, n_states = NA_integer_, note = conditionMessage(out), url = u)
  } else {
    tibble::tibble(year = y, ok = nrow(out) >= 45, n_states = nrow(out), note = "ok", url = u)
  }
})

print(probe, n = Inf)

good_years <- probe |> dplyr::filter(ok) |> dplyr::pull(year)
cat("\nEarliest good year:", ifelse(length(good_years) > 0, min(good_years), NA), "\n")
```
