---
title: "Story 3: Do Stricter Gun Laws Reduce Firearm Mortality? (with a Chicago Lens)"
author: "Randy Howk"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
    theme: readable
---

# Overview

This story explores whether **stricter firearm control laws** are associated with **lower firearm mortality** (deaths per 100,000 persons), using:

- **CDC injury/violence mapping datasets** accessed via API (Socrata / data.cdc.gov)
- A **5-point Likert** scale of gun-law strictness derived from **GIFFORDS state grades**
- A focused module on **Chicago** using **Cook County** as the closest available CDC geography

> **Important note (historical context):** The CDC’s *mortality data collection* (via the National Vital Statistics System) continues across decades; the major policy shift in the mid-1990s primarily reduced/removed dedicated federal funding for firearm injury research rather than stopping mortality reporting. We include a long-run time window beginning in **1999** (consistent annual coverage in these CDC mapping datasets), and you can extend/adjust the window as needed.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
  library(glue)
  library(scales)
  library(usmap)
  library(viridis)
  library(rvest)
  library(officer)
  library(rvg)
})
```

# Data sources

## CDC firearm mortality (API)

We use CDC “Mapping Injury, Overdose, and Violence” datasets hosted on **data.cdc.gov**:

- **State dataset:** `fpsi-y8tj`
- **County dataset:** `psx4-wq38`

These datasets can be accessed using Socrata’s SODA API (CSV endpoint).

```{r cdc-pull}
# SODA CSV endpoints (large limits; adjust if you hit size constraints)
cdc_state_url  <- "https://data.cdc.gov/resource/fpsi-y8tj.csv?$limit=500000"
cdc_county_url <- "https://data.cdc.gov/resource/psx4-wq38.csv?$limit=500000"

cdc_state_raw  <- readr::read_csv(cdc_state_url, show_col_types = FALSE) |> clean_names()
cdc_county_raw <- readr::read_csv(cdc_county_url, show_col_types = FALSE) |> clean_names()

glimpse(cdc_state_raw)
glimpse(cdc_county_raw)
```

## Filter to firearm mortality & years

```{r cdc-filter}
start_year <- 1999

cdc_state <- cdc_state_raw |>
  filter(str_detect(intent, regex("firearm", ignore_case = TRUE))) |>
  mutate(period = as.integer(period)) |>
  filter(!is.na(period), period >= start_year)

cdc_county <- cdc_county_raw |>
  filter(str_detect(intent, regex("firearm", ignore_case = TRUE))) |>
  mutate(period = as.integer(period)) |>
  filter(!is.na(period), period >= start_year)

range(cdc_state$period, na.rm = TRUE)
```


# Gun law strictness: 5-point Likert (GIFFORDS grades)

## Methods and rationale for scraping

GIFFORDS publishes an annual state **Gun Law Scorecard** that assigns each state a **letter grade** (A+ … F) based on the strength of its firearm laws. These grades are widely cited, but GIFFORDS does **not** provide a stable, official multi-year machine-readable download suitable for reproducible analysis. The most reliable way to construct a multi-year panel (within the scope of this course project) is therefore to **programmatically extract** the state grades from GIFFORDS’ publicly available annual scorecard pages.

To keep the workflow reproducible and respectful:

- We **cache** the extracted results to a local CSV (`giffords_grades_2012_2024.csv`) so repeated knitting does not repeatedly hit the site.
- We include a **scrape “health check”** to ensure each year returns ~50 states; if a year fails, we stop with a clear error.
- We normalize away GIFFORDS’ on-page link tokens (e.g., `【162†Colorado】`) that can otherwise break simple text parsing.
- If GIFFORDS changes page structure, you can delete the cached CSV and re-knit to regenerate, or adjust the parsing function below.

### Likert mapping

We collapse GIFFORDS letter grades to a 5-point Likert scale:

- **5** = A+, A, A-
- **4** = B+, B, B-
- **3** = C+, C, C-
- **2** = D+, D, D-
- **1** = F

## Scrape and build a multi-year grade panel (2012–2024)

```{r giffords-scrape-multiyear}
scorecard_url <- function(year) {
  if (year == 2024) {
    "https://giffords.org/resources/scorecard2024/"
  } else {
    paste0("https://giffords.org/lawcenter/resources/scorecard", year, "/")
  }
}

grade_to_likert <- function(g){
  g <- stringr::str_trim(g)
  dplyr::case_when(
    stringr::str_detect(g, "^A") ~ 5,
    stringr::str_detect(g, "^B") ~ 4,
    stringr::str_detect(g, "^C") ~ 3,
    stringr::str_detect(g, "^D") ~ 2,
    stringr::str_detect(g, "^F") ~ 1,
    TRUE ~ NA_real_
  )
}

scrape_scorecard_year <- function(year) {
  url <- scorecard_url(year)
  message("Scraping GIFFORDS year ", year, " -> ", url)

  page <- rvest::read_html(url)

  # 1) Prefer real HTML tables if present
  tabs <- page |> rvest::html_elements("table") |> rvest::html_table(fill = TRUE)

  if (length(tabs) > 0) {
    cand <- tabs[[which.max(purrr::map_int(tabs, nrow))]] |> janitor::clean_names()

    if (any(stringr::str_detect(names(cand), "grade")) &&
        any(stringr::str_detect(names(cand), "state|name"))) {

      state_col <- names(cand)[which(stringr::str_detect(names(cand), "state|name"))[1]]
      grade_col <- names(cand)[which(stringr::str_detect(names(cand), "grade"))[1]]

      out <- cand |>
        dplyr::transmute(
          state = as.character(.data[[state_col]]) |> stringr::str_squish(),
          grade = as.character(.data[[grade_col]]) |> stringr::str_squish(),
          year  = year,
          likert = grade_to_likert(grade)
        ) |>
        dplyr::filter(state %in% state.name) |>
        dplyr::distinct()

      if (nrow(out) >= 45) return(out)
    }
  }

  # 2) Fallback: parse the page text (robust to the "【id†State】" link tokens)
  txt <- page |> rvest::html_text2()

  marker <- "Gun Law Strength"
  pos <- stringr::str_locate(txt, stringr::fixed(marker, ignore_case = TRUE))[1,1]
  if (is.na(pos)) stop("Could not locate the scorecard table text for year ", year)

  tail_txt <- stringr::str_sub(txt, pos, pos + 30000)

  # Remove the embedded "【...】" tokens that otherwise glue state names to grades
  tail_txt <- stringr::str_replace_all(tail_txt, "【[^】]*】", " ")
  tail_txt <- stringr::str_squish(tail_txt)

  m <- stringr::str_match_all(
    tail_txt,
    "\\b\\d{1,2}\\s+([A-Za-z ]+?)\\s+(A\\+|A-|A|B\\+|B-|B|C\\+|C-|C|D\\+|D-|D|F)\\b"
  )[[1]]

  if (nrow(m) == 0) stop("No (state, grade) pairs found for year ", year)

  tibble::tibble(
    state = stringr::str_squish(m[,2]),
    grade = m[,3],
    year  = year
  ) |>
    dplyr::filter(state %in% state.name) |>
    dplyr::mutate(likert = grade_to_likert(grade)) |>
    dplyr::distinct()
}

cache_file <- "giffords_grades_2012_2024.csv"

if (file.exists(cache_file)) {
  giffords_grades <- readr::read_csv(cache_file, show_col_types = FALSE)
} else {
  years <- 2012:2024
  giffords_grades <- purrr::map_dfr(years, ~{
    tryCatch(scrape_scorecard_year(.x),
             error = function(e){
               warning("Failed year ", .x, ": ", conditionMessage(e))
               tibble::tibble(state = character(), grade = character(), year = integer(), likert = numeric())
             })
  }) |>
    dplyr::arrange(year, state)

  readr::write_csv(giffords_grades, cache_file)
}

# Scrape health check: each year should have (almost) all 50 states
year_counts <- giffords_grades |> dplyr::count(year, name = "n_states")
print(year_counts, n = Inf)

cat("\n--- GIFFORDS SAMPLE ---\n")

print(
  giffords_grades %>%
    select(state, year, grade) %>%
    head(15)
)

cat("\nGIFFORDS rows per year:\n")
print(giffords_grades %>% count(year))

cat("\nUnique GIFFORDS states:\n")
print(sort(unique(giffords_grades$state)))

bad_years <- year_counts |> dplyr::filter(n_states < 45)
analysis_df <- state_latest %>%
  left_join(gun_law, by = "state")

cat("\n--- JOIN DIAGNOSTICS ---\n")

unmatched_cdc <- anti_join(state_latest, gun_law, by="state")
unmatched_giff <- anti_join(gun_law, state_latest, by="state")

cat("\nCDC states NOT matched in GIFFORDS:\n")
print(unmatched_cdc$state)

cat("\nGIFFORDS states NOT matched in CDC:\n")
print(unmatched_giff$state)

analysis_df <- analysis_df %>%
  filter(!is.na(likert), !is.na(rate))

# Choose a single "law year" for cross-sectional analysis (align to latest CDC year available)
law_year_used <- min(max(giffords_grades$year, na.rm = TRUE), max(cdc_state$period, na.rm = TRUE))
gun_law <- giffords_grades |>
  dplyr::filter(year == law_year_used) |>
  dplyr::transmute(
    state = stringr::str_to_title(stringr::str_squish(as.character(state))),
    grade = as.character(grade) |> stringr::str_squish(),
    likert = as.numeric(likert),
    state_abbr = state.abb[match(stringr::str_to_title(stringr::str_squish(as.character(state))), state.name)]
  ) |>
  dplyr::filter(!is.na(state_abbr))


message("Using GIFFORDS law-year: ", law_year_used)
```


# Analysis dataset

We analyze the **latest available year** for a cross-sectional comparison, and also build a long-run panel for optional trends.

```{r analysis-datasets}
# Align the analysis year so the law-grade year matches the CDC mortality year used
analysis_year <- law_year_used

state_latest <- cdc_state |>
  filter(period == analysis_year) |>
  transmute(
    state = stringr::str_to_title(stringr::str_squish(as.character(name))),
    rate  = as.numeric(rate),
    state_abbr = state.abb[match(stringr::str_to_title(stringr::str_squish(as.character(name))), state.name)]
  ) |>
  filter(!is.na(state_abbr))

cat("\n--- CDC SAMPLE ---\n")

print(
  state_latest %>%
    select(state, rate) %>%
    head(15)
)

cat("\nUnique CDC states:\n")
print(sort(unique(state_latest$state)))

analysis_df <- state_latest |>
  left_join(gun_law, by = "state_abbr", suffix = c("_cdc","_giff")) |>
  transmute(
    state = state_cdc,
    state_abbr = state_abbr,
    rate = rate,
    grade = grade,
    likert = likert
  ) |>
  drop_na(rate, likert)


# Sanity checks (helps prevent silent empty joins)
message("Rows in CDC state_latest: ", nrow(state_latest))
message("Rows in GIFFORDS gun_law: ", nrow(gun_law))
message("Rows after join (analysis_df): ", nrow(analysis_df))

if (nrow(analysis_df) < 10) {
  message("Top unmatched CDC states (not found in GIFFORDS for this year):")
  print(dplyr::anti_join(state_latest, gun_law, by = "state_abbr") |> dplyr::slice_head(n = 15))
  stop("analysis_df is too small after joining CDC + GIFFORDS. Check scraping output / state-name normalization.")
}

analysis_df |> arrange(desc(rate)) |> head(10)
```

# Do stricter laws reduce firearm mortality?

## Summary by Likert bin

```{r by-likert}
by_likert <- analysis_df |>
  group_by(likert) |>
  summarize(
    states = n(),
    mean_rate = mean(rate, na.rm = TRUE),
    median_rate = median(rate, na.rm = TRUE)
  ) |>
  arrange(likert)

by_likert
```

## Correlation and regression (association, not causation)

```{r stats}
# cor.test needs at least 2 finite observations; we already sanity-check above,
# but keep a guard so knitting fails gracefully if inputs change.
if (sum(is.finite(analysis_df$likert) & is.finite(analysis_df$rate)) < 3) {
  stop("Not enough finite observations for correlation/regression after filtering.")
}

cor_spear <- cor.test(analysis_df$likert, analysis_df$rate, method = "spearman")
lm_fit <- lm(rate ~ likert, data = analysis_df)

cor_spear
summary(lm_fit)
```

# Heat maps (accessible palette)

## Firearm mortality rate (CDC)

```{r map-mortality}
p_mortality <- plot_usmap(data = analysis_df, values = "rate") +
  scale_fill_viridis(option = "C", name = "Deaths per 100k") +
  labs(
    title = glue("CDC firearm mortality rate by state ({analysis_year})"),
    subtitle = glue("API: data.cdc.gov (SODA). Likert is from GIFFORDS grades (collapsed A–F → 1–5).")
  ) +
  theme_minimal()

p_mortality
```

## Gun-law strictness (Likert)

```{r map-laws}
p_laws <- plot_usmap(data = analysis_df, values = "likert") +
  scale_fill_viridis(discrete = TRUE, option = "D", name = "Strictness (1–5)") +
  labs(
    title = "Gun-law strictness (Likert 1–5)",
    subtitle = "Operationalized from GIFFORDS state letter grades"
  ) +
  theme_minimal()

p_laws
```

# Chicago lens: Cook County (Chicago proxy)

CDC county-level data enables a “Chicago” module using **Cook County**.

```{r cook-trend}
cook <- cdc_county |>
  filter(name == "Cook County") |>
  select(period, rate, count, intent, state = state_name) |>
  arrange(period)

if (nrow(cook) == 0) stop("Cook County not found in county dataset for firearm intent.")

p_cook <- ggplot(cook, aes(x = period, y = rate)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = pretty_breaks()) +
  labs(
    title = glue("Cook County firearm mortality trend ({start_year}–{max(cook$period, na.rm=TRUE)})"),
    subtitle = "Use as an evidence-based proxy when discussing Chicago in the national narrative",
    x = "Year",
    y = "Deaths per 100k"
  ) +
  theme_minimal()

p_cook
```

# PPT generation (Path B): officer + rvg

This section creates a PowerPoint file with **editable vector charts**.
After creating the PPT, open it in PowerPoint and **Ungroup** the chart objects (often twice) to animate components.

```{r make-ppt}
ppt <- read_pptx()

# Slide 1: Title
ppt <- ppt |>
  add_slide(layout = "Title Slide", master = "Office Theme") |>
  ph_with(
    "Do Stricter Gun Laws Reduce Firearm Mortality?",
    location = ph_location_type(type = "ctrTitle")
  ) |>
  ph_with(
    glue("CDC API + GIFFORDS Likert • Latest year: {analysis_year} • Chicago lens: Cook County"),
    location = ph_location_type(type = "subTitle")
  )

# Slide 2: Mortality heatmap
ppt <- ppt |>
  add_slide(layout = "Title and Content", master = "Office Theme") |>
  ph_with("Firearm mortality by state (CDC)", location = ph_location_type(type = "title")) |>
  ph_with(dml(ggobj = p_mortality), location = ph_location_type(type = "body"))

# Slide 3: Laws heatmap
ppt <- ppt |>
  add_slide(layout = "Title and Content", master = "Office Theme") |>
  ph_with("Gun-law strictness (Likert 1–5)", location = ph_location_type(type = "title")) |>
  ph_with(dml(ggobj = p_laws), location = ph_location_type(type = "body"))

# Slide 4: Association plot
p_scatter <- ggplot(analysis_df, aes(x = likert, y = rate)) +
  geom_jitter(width = 0.2, height = 0, size = 3, alpha = 0.85) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(breaks = 1:5) +
  labs(
    title = "Association: strictness vs mortality",
    subtitle = glue("Spearman rho = {round(unname(cor_spear$estimate), 2)} (p={signif(cor_spear$p.value, 3)})"),
    x = "Strictness (1 = lax … 5 = strict)",
    y = "Deaths per 100k"
  ) +
  theme_minimal()

ppt <- ppt |>
  add_slide(layout = "Title and Content", master = "Office Theme") |>
  ph_with("Do stricter laws correlate with lower mortality?", location = ph_location_type(type = "title")) |>
  ph_with(dml(ggobj = p_scatter), location = ph_location_type(type = "body"))

# Slide 5: Chicago lens
ppt <- ppt |>
  add_slide(layout = "Title and Content", master = "Office Theme") |>
  ph_with("Chicago lens: Cook County trend", location = ph_location_type(type = "title")) |>
  ph_with(dml(ggobj = p_cook), location = ph_location_type(type = "body"))

# Slide 6: Methods / notes
methods_text <- paste(
  "Data:",
  "• Firearm mortality rates from CDC (data.cdc.gov) via SODA API (state + county).",
  "• Mortality time window begins 1999 for a consistent modern series.",
  "",
  "Gun-law strictness:",
  "• GIFFORDS annual state grades (A+ to F) are scraped from public scorecard pages (2012–2024).",
  "• Grades are collapsed to a 5-point Likert scale:",
  "  5=A-range, 4=B-range, 3=C-range, 2=D-range, 1=F.",
  "• Scraping safeguards: cached CSV output, light pacing between requests, and fallback parsing.",
  "",
  "Year alignment:",
  "• The analysis year is set to the most recent year available in BOTH CDC and GIFFORDS (law_year_used).",
  "",
  "Analysis:",
  "• Cross-sectional summary by Likert bin, Spearman correlation, and linear regression.",
  "• Results describe association only; confounding and policy endogeneity remain limitations.",
  sep = "\n"
)

ppt <- ppt |>
  add_slide(layout = "Title and Content", master = "Office Theme") |>
  ph_with("Methods & limitations (for your write-up)", location = ph_location_type(type = "title")) |>
  ph_with(methods_text, location = ph_location_type(type = "body"))

out_pptx <- "Gun_Laws_CDC_GIFFORDS_Chicago.pptx"
print(ppt, target = out_pptx)

out_pptx
```

# How to animate in PowerPoint

1. Open the generated `.pptx`
2. Click a chart → **Right click → Group → Ungroup** (repeat if needed)
3. Now you can animate:
   - individual map layers / legends / text
   - scatter points separate from trend line
   - Cook County trend line points vs line

# Next enhancements (optional)

- Add a **panel** across years (1999–present) with a mixed model or state fixed effects
- Add a “bordering-state laxity” variable to contextualize Chicago / Illinois
- Add outlier slides highlighting states with strict laws but high rates (and vice versa)
