---
title: "Story 3: Do Stricter Gun Laws Reduce Firearm Mortality?"
author: "Randy Howk"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
    theme: readable
---

# Overview

This story explores whether **stricter firearm control laws** are associated with **lower firearm mortality** (deaths per 100,000 persons), using:

- **CDC injury/violence mapping datasets** accessed via API (Socrata / data.cdc.gov)
- A **5-point Likert** scale of gun-law strictness derived from **GIFFORDS state grades**
- State-level CDC mortality plus GIFFORDS strictness grading

> **Important note (historical context):** The CDC’s *mortality data collection* (via the National Vital Statistics System) continues across decades; the major policy shift in the mid-1990s primarily reduced/removed dedicated federal funding for firearm injury research rather than stopping mortality reporting. We include a long-run time window beginning in **1999** (consistent annual coverage in these CDC mapping datasets), and you can extend/adjust the window as needed.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
  library(glue)
  library(scales)
  library(usmap)
  library(viridis)
  library(rvest)
  library(officer)
  library(rvg)
})

# Guardrail for IDE runs: render this Rmd, not CSV debug artifacts.
current_input <- tryCatch(basename(knitr::current_input()), error = function(e) "")
if (nzchar(current_input) && current_input != "Story3.rmd") {
  message(
    "This document is intended to run from Story3.rmd. ",
    "Do not render CSV debug files directly."
  )
}

# Normalize state labels from mixed sources to 2-letter USPS abbreviations.
# Handles full state names, plain abbreviations, and labels like "Alabama (AL)".
to_state_abbr <- function(x) {
  x <- as.character(x)
  x <- stringr::str_squish(x)
  x <- stringr::str_to_title(x)

  # If already a plain 2-letter code, keep it.
  is_code <- stringr::str_detect(x, "^[A-Za-z]{2}$")
  out <- rep(NA_character_, length(x))
  out[is_code] <- stringr::str_to_upper(x[is_code])

  # Parse labels with trailing "(XX)" code.
  paren_code <- stringr::str_match(x, "\\(([A-Za-z]{2})\\)$")[, 2]
  out[is.na(out) & !is.na(paren_code)] <- stringr::str_to_upper(paren_code[is.na(out) & !is.na(paren_code)])

  # Strip any parenthetical suffix and map by full name.
  base_name <- stringr::str_squish(stringr::str_replace(x, "\\s*\\(.*\\)$", ""))
  idx <- match(base_name, state.name)
  out[is.na(out) & !is.na(idx)] <- state.abb[idx[is.na(out) & !is.na(idx)]]

  # Keep only true US states for this project.
  out[!out %in% state.abb] <- NA_character_
  out
}
```

# Data sources

## CDC firearm mortality (API)

We use CDC “Mapping Injury, Overdose, and Violence” datasets hosted on **data.cdc.gov**:

- **State dataset:** `fpsi-y8tj`
- **County dataset:** `psx4-wq38`

These datasets can be accessed using Socrata’s SODA API (CSV endpoint).

```{r cdc-pull}
# SODA CSV endpoints (large limits; adjust if you hit size constraints)
cdc_state_url  <- "https://data.cdc.gov/resource/fpsi-y8tj.csv?$limit=500000"
cdc_county_url <- "https://data.cdc.gov/resource/psx4-wq38.csv?$limit=500000"

cdc_state_raw  <- readr::read_csv(cdc_state_url, show_col_types = FALSE) |> clean_names()
cdc_county_raw <- readr::read_csv(cdc_county_url, show_col_types = FALSE) |> clean_names()

glimpse(cdc_state_raw)
glimpse(cdc_county_raw)

message("CDC state raw rows: ", nrow(cdc_state_raw), " | cols: ", ncol(cdc_state_raw))
message("CDC county raw rows: ", nrow(cdc_county_raw), " | cols: ", ncol(cdc_county_raw))

# Write small raw snapshots for debugging when downstream filters go empty.
readr::write_csv(utils::head(cdc_state_raw, 200), "cdc_state_raw_sample.csv")
readr::write_csv(utils::head(cdc_county_raw, 200), "cdc_county_raw_sample.csv")
```

## Filter to firearm mortality & years

```{r cdc-filter}
start_year <- 1999

state_intents <- sort(unique(as.character(cdc_state_raw$intent)))
county_intents <- sort(unique(as.character(cdc_county_raw$intent)))

message("Sample CDC state intents: ", paste(head(state_intents, 10), collapse = ", "))
message("Sample CDC county intents: ", paste(head(county_intents, 10), collapse = ", "))

# Prefer current CDC schema (FA_Deaths + FA_Homicide); fall back to older text-based intent labels.
has_fa_deaths_state <- "FA_Deaths" %in% state_intents
has_fa_deaths_county <- "FA_Deaths" %in% county_intents
has_fa_homicide_state <- "FA_Homicide" %in% state_intents
has_fa_homicide_county <- "FA_Homicide" %in% county_intents

if (has_fa_deaths_state || has_fa_homicide_state) {
  state_keep <- intersect(c("FA_Deaths", "FA_Homicide"), state_intents)
  cdc_state <- cdc_state_raw |>
    filter(intent %in% state_keep)
  message("Using CDC state intents: ", paste(state_keep, collapse = ", "))
} else {
  cdc_state <- cdc_state_raw |>
    filter(str_detect(intent, regex("firearm", ignore_case = TRUE)))
  message("Using CDC state fallback intent filter: text match on 'firearm'")
}

if (has_fa_deaths_county || has_fa_homicide_county) {
  county_keep <- intersect(c("FA_Deaths", "FA_Homicide"), county_intents)
  cdc_county <- cdc_county_raw |>
    filter(intent %in% county_keep)
  message("Using CDC county intents: ", paste(county_keep, collapse = ", "))
} else {
  cdc_county <- cdc_county_raw |>
    filter(str_detect(intent, regex("firearm", ignore_case = TRUE)))
  message("Using CDC county fallback intent filter: text match on 'firearm'")
}

cdc_state <- cdc_state |>
  mutate(period = suppressWarnings(as.integer(period))) |>
  filter(!is.na(period), period >= start_year)

cdc_county <- cdc_county |>
  mutate(period = suppressWarnings(as.integer(period))) |>
  filter(!is.na(period), period >= start_year)

message("CDC state rows after firearm/year filter: ", nrow(cdc_state))
message("CDC county rows after firearm/year filter: ", nrow(cdc_county))

if (nrow(cdc_state) == 0) {
  readr::write_csv(utils::head(cdc_state_raw, 500), "cdc_state_filter_debug.csv")
  stop(
    "CDC state data is empty after filter. ",
    "Check cdc_state_raw_sample.csv and cdc_state_filter_debug.csv ",
    "for available columns/values (especially `intent`)."
  )
}

range(cdc_state$period, na.rm = TRUE)
```


# Gun law strictness: 5-point Likert (GIFFORDS grades)

## Methods and rationale for scraping

GIFFORDS publishes an annual state **Gun Law Scorecard** that assigns each state a **letter grade** (A+ … F) based on the strength of its firearm laws. These grades are widely cited, but GIFFORDS does **not** provide a stable, official multi-year machine-readable download suitable for reproducible analysis. The most reliable way to construct a multi-year panel (within the scope of this course project) is therefore to **programmatically extract** the state grades from GIFFORDS’ publicly available annual scorecard pages.

To keep the workflow reproducible and respectful:

- We **cache** the extracted results to a local CSV (`giffords_grades_2012_2024.csv`) so repeated knitting does not repeatedly hit the site.
- We include a **scrape “health check”** to ensure each year returns ~50 states; if a year fails, we stop with a clear error.
- We normalize away GIFFORDS’ on-page link tokens (e.g., `【162†Colorado】`) that can otherwise break simple text parsing.
- If GIFFORDS changes page structure, you can delete the cached CSV and re-knit to regenerate, or adjust the parsing function below.

### Likert mapping

We collapse GIFFORDS letter grades to a 5-point Likert scale:

- **5** = A+, A, A-
- **4** = B+, B, B-
- **3** = C+, C, C-
- **2** = D+, D, D-
- **1** = F

## Scrape and build a multi-year grade panel (2012–2024)

```{r giffords-scrape-multiyear}
scorecard_url <- function(year) {
  if (year == 2024) {
    "https://giffords.org/resources/scorecard2024/"
  } else {
    paste0("https://giffords.org/lawcenter/resources/scorecard", year, "/")
  }
}

grade_to_likert <- function(g){
  g <- stringr::str_trim(g)
  dplyr::case_when(
    stringr::str_detect(g, "^A") ~ 5,
    stringr::str_detect(g, "^B") ~ 4,
    stringr::str_detect(g, "^C") ~ 3,
    stringr::str_detect(g, "^D") ~ 2,
    stringr::str_detect(g, "^F") ~ 1,
    TRUE ~ NA_real_
  )
}

scrape_scorecard_year <- function(year) {
  url <- scorecard_url(year)
  message("Scraping GIFFORDS year ", year, " -> ", url)

  page <- rvest::read_html(url)

  # 1) Prefer real HTML tables if present
  tabs <- page |> rvest::html_elements("table") |> rvest::html_table(fill = TRUE)

  if (length(tabs) > 0) {
    cand <- tabs[[which.max(purrr::map_int(tabs, nrow))]] |> janitor::clean_names()

    if (any(stringr::str_detect(names(cand), "grade")) &&
        any(stringr::str_detect(names(cand), "state|name"))) {

      state_col <- names(cand)[which(stringr::str_detect(names(cand), "state|name"))[1]]
      grade_col <- names(cand)[which(stringr::str_detect(names(cand), "grade"))[1]]

      out <- cand |>
        dplyr::transmute(
          state = as.character(.data[[state_col]]) |> stringr::str_squish(),
          grade = as.character(.data[[grade_col]]) |> stringr::str_squish(),
          year  = year,
          likert = grade_to_likert(grade)
        ) |>
        dplyr::filter(state %in% state.name) |>
        dplyr::distinct()

      if (nrow(out) >= 45) return(out)
    }
  }

  # 2) Fallback: parse the page text (robust to the "【id†State】" link tokens)
  txt <- page |> rvest::html_text2()

  marker <- "Gun Law Strength"
  pos <- stringr::str_locate(txt, stringr::fixed(marker, ignore_case = TRUE))[1,1]
  if (is.na(pos)) stop("Could not locate the scorecard table text for year ", year)

  tail_txt <- stringr::str_sub(txt, pos, pos + 30000)

  # Remove the embedded "【...】" tokens that otherwise glue state names to grades
  tail_txt <- stringr::str_replace_all(tail_txt, "【[^】]*】", " ")
  tail_txt <- stringr::str_squish(tail_txt)

  m <- stringr::str_match_all(
    tail_txt,
    "\\b\\d{1,2}\\s+([A-Za-z ]+?)\\s+(A\\+|A-|A|B\\+|B-|B|C\\+|C-|C|D\\+|D-|D|F)\\b"
  )[[1]]

  if (nrow(m) == 0) stop("No (state, grade) pairs found for year ", year)

  tibble::tibble(
    state = stringr::str_squish(m[,2]),
    grade = m[,3],
    year  = year
  ) |>
    dplyr::filter(state %in% state.name) |>
    dplyr::mutate(likert = grade_to_likert(grade)) |>
    dplyr::distinct()
}

cache_file <- "giffords_grades_2012_2024.csv"

if (file.exists(cache_file)) {
  giffords_grades <- readr::read_csv(cache_file, show_col_types = FALSE)
} else {
  years <- 2012:2024
  giffords_grades <- purrr::map_dfr(years, ~{
    tryCatch(scrape_scorecard_year(.x),
             error = function(e){
               warning("Failed year ", .x, ": ", conditionMessage(e))
               tibble::tibble(state = character(), grade = character(), year = integer(), likert = numeric())
             })
  }) |>
    dplyr::arrange(year, state)

  readr::write_csv(giffords_grades, cache_file)
}

# Scrape health check: each year should have (almost) all 50 states
year_counts <- giffords_grades |> dplyr::count(year, name = "n_states")
print(year_counts, n = Inf)

cat("\n--- GIFFORDS SAMPLE ---\n")

print(
  giffords_grades %>%
    select(state, year, grade) %>%
    head(15)
)

cat("\nGIFFORDS rows per year:\n")
print(giffords_grades %>% count(year))

cat("\nUnique GIFFORDS states:\n")
print(sort(unique(giffords_grades$state)))

bad_years <- year_counts |> dplyr::filter(n_states < 45)

if (nrow(bad_years) > 0) {
  warning(
    "Some GIFFORDS years returned fewer than 45 states: ",
    paste(bad_years$year, collapse = ", ")
  )
}

# Choose a single analysis year using overlap + CDC coverage.
# This avoids selecting a very recent partial CDC year with too few states.
cdc_debug <- cdc_state |>
  dplyr::transmute(
    period_raw = as.character(period),
    period = suppressWarnings(as.integer(period)),
    name = as.character(name),
    state_abbr = to_state_abbr(name)
  )

giffords_debug <- giffords_grades |>
  dplyr::transmute(
    state = as.character(state),
    year = suppressWarnings(as.integer(year)),
    grade = as.character(grade),
    likert = as.numeric(likert),
    state_abbr = to_state_abbr(state)
  )

# Debug exports to inspect matching/years when knitting fails
# Note: these are diagnostics only. Render Story3.rmd, not cdc.csv/giffords.csv.
readr::write_csv(cdc_debug, "cdc.csv")
readr::write_csv(giffords_debug, "giffords.csv")

cdc_year_coverage <- cdc_debug |>
  dplyr::select(period, state_abbr) |>
  dplyr::filter(!is.na(period), !is.na(state_abbr)) |>
  dplyr::distinct(period, state_abbr) |>
  dplyr::count(period, name = "n_states") |>
  dplyr::arrange(desc(period))

cat("\nCDC state coverage by year (top 10):\n")
print(head(cdc_year_coverage, 10))

giffords_years <- sort(unique(giffords_debug$year[!is.na(giffords_debug$year)]))
common_years <- sort(intersect(giffords_years, cdc_year_coverage$period))

overlap_debug <- tibble::tibble(
  cdc_years = paste(sort(unique(cdc_year_coverage$period)), collapse = ";"),
  giffords_years = paste(sort(unique(giffords_years)), collapse = ";"),
  common_years = paste(common_years, collapse = ";")
)
readr::write_csv(overlap_debug, "overlap_debug.csv")

if (length(common_years) == 0) {
  stop(
    "No overlapping year found between CDC and GIFFORDS. ",
    "Wrote debug files: cdc.csv, giffords.csv, overlap_debug.csv"
  )
}

year_quality <- tibble::tibble(year = common_years) |>
  dplyr::left_join(
    cdc_year_coverage |>
      dplyr::transmute(year = period, cdc_states = n_states),
    by = "year"
  ) |>
  dplyr::mutate(score = cdc_states)

cat("\nOverlapping CDC/GIFFORDS years:\n")
print(year_quality |> dplyr::arrange(desc(year)))

law_year_used <- year_quality |>
  dplyr::arrange(desc(score), desc(year)) |>
  dplyr::slice(1) |>
  dplyr::pull(year)
gun_law <- giffords_grades |>
  dplyr::filter(year == law_year_used) |>
  dplyr::transmute(
    state = stringr::str_to_title(stringr::str_squish(as.character(state))),
    grade = as.character(grade) |> stringr::str_squish(),
    likert = as.numeric(likert),
    state_abbr = to_state_abbr(state)
  ) |>
  dplyr::filter(!is.na(state_abbr))


message("Using GIFFORDS law-year: ", law_year_used)
```


# Analysis dataset

We analyze the **latest available year** for a cross-sectional comparison, and also build a long-run panel for optional trends.

```{r analysis-datasets}
# Align the analysis year so the law-grade year matches the CDC mortality year used
analysis_year <- law_year_used

state_latest <- cdc_state |>
  filter(period == analysis_year) |>
  transmute(
    state = stringr::str_to_title(stringr::str_squish(as.character(name))),
    rate  = as.numeric(rate),
    state_abbr = to_state_abbr(name)
  ) |>
  filter(!is.na(state_abbr), !is.na(rate)) |>
  group_by(state_abbr) |>
  summarize(
    state = first(state),
    # Combined-intent proxy (FA_Deaths + FA_Homicide), not additive mortality.
    rate = mean(rate, na.rm = TRUE),
    .groups = "drop"
  )

cat("\n--- CDC SAMPLE ---\n")

print(
  state_latest %>%
    select(state, rate) %>%
    head(15)
)

cat("\nUnique CDC states:\n")
print(sort(unique(state_latest$state)))

analysis_df <- state_latest |>
  left_join(gun_law, by = "state_abbr", suffix = c("_cdc","_giff")) |>
  transmute(
    state = state_cdc,
    state_abbr = state_abbr,
    rate_per_100k = rate,
    grade = grade,
    likert = likert
  ) |>
  drop_na(rate_per_100k, likert) |>
  mutate(
    rate_per_capita = rate_per_100k / 100000,
    law_band = factor(likert, levels = 1:5, labels = c("F", "D", "C", "B", "A"), ordered = TRUE),
    strictness_group = dplyr::case_when(
      likert %in% c(4, 5) ~ "Strict (4-5)",
      likert %in% c(1, 2, 3) ~ "Less strict (1-3)",
      TRUE ~ NA_character_
    )
  ) |>
  filter(state_abbr != "PR")


# Sanity checks (helps prevent silent empty joins)
message("Rows in CDC state_latest: ", nrow(state_latest))
message("Rows in GIFFORDS gun_law: ", nrow(gun_law))
message("Rows after join (analysis_df): ", nrow(analysis_df))

if (nrow(analysis_df) < 10) {
  message("Top unmatched CDC states (not found in GIFFORDS for this year):")
  print(dplyr::anti_join(state_latest, gun_law, by = "state_abbr") |> dplyr::slice_head(n = 15))
  stop("analysis_df is too small after joining CDC + GIFFORDS. Check scraping output / state-name normalization.")
}

analysis_df |> arrange(desc(rate_per_100k)) |> head(10)
```

# Do stricter laws reduce firearm mortality?

## Summary by Likert bin

```{r by-likert}
by_likert <- analysis_df |>
  group_by(likert) |>
  summarize(
    states = n(),
    mean_rate_per_100k = mean(rate_per_100k, na.rm = TRUE),
    median_rate_per_100k = median(rate_per_100k, na.rm = TRUE),
    mean_rate_per_capita = mean(rate_per_capita, na.rm = TRUE),
    median_rate_per_capita = median(rate_per_capita, na.rm = TRUE)
  ) |>
  arrange(likert)

by_likert
```

## Combined groups comparison: strict (4–5) vs less strict (1–3)

```{r strict-vs-less}
comparison_45_vs_123 <- analysis_df |>
  filter(!is.na(strictness_group)) |>
  group_by(strictness_group) |>
  summarize(
    states = n(),
    mean_rate_per_100k = mean(rate_per_100k, na.rm = TRUE),
    median_rate_per_100k = median(rate_per_100k, na.rm = TRUE),
    mean_rate_per_capita = mean(rate_per_capita, na.rm = TRUE),
    median_rate_per_capita = median(rate_per_capita, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(strictness_group = factor(strictness_group, levels = c("Strict (4-5)", "Less strict (1-3)"))) |>
  arrange(strictness_group)

comparison_45_vs_123
```

## Heat maps (accessible palette)

## Gun-law strictness (Likert)

```{r map-laws}
p_laws <- plot_usmap(data = analysis_df, values = "law_band") +
  scale_fill_manual(
    values = c(
      "F" = "#c62828",    # least strict
      "D" = "#f4b6c2",
      "C" = "#d9d9d9",
      "B" = "#7fb3ff",
      "A" = "#0b3c8c"     # strictest
    ),
    drop = FALSE,
    na.translate = FALSE,
    name = "Law grade band"
  ) +
  labs(
    title = "Gun-law strictness map (F red → A blue)",
    subtitle = "A=blue, B=light blue, C=light gray, D=light pink, F=red"
  ) +
  theme_void() +
  theme(legend.position = "right")

p_laws
```

## Firearm mortality rate (CDC)

```{r map-mortality}
p_mortality <- plot_usmap(data = analysis_df, values = "rate_per_100k") +
  scale_fill_gradient(
    low = "#fff7bc",  # fewest deaths (light yellow)
    high = "#b10026", # most deaths (red)
    name = "Deaths per 100k"
  ) +
  labs(
    title = glue("CDC firearm mortality rate by state ({analysis_year})"),
    subtitle = glue("API: data.cdc.gov (SODA). Likert is from GIFFORDS grades (collapsed A–F → 1–5).")
  ) +
  theme_void() +
  theme(legend.position = "right")

p_mortality
```

## Combined map: strictness color + mortality distance scaling

```{r map-combined-scaled}
map_weighted_mean <- mean(analysis_df$rate_per_100k, na.rm = TRUE)

state_sf <- usmap::us_map(regions = "states")
state_coords <- as.data.frame(sf::st_coordinates(state_sf$geom))
state_attr <- state_sf |>
  sf::st_drop_geometry() |>
  dplyr::mutate(row_id = dplyr::row_number())

state_shapes <- state_coords |>
  dplyr::transmute(
    row_id = L3,
    group = paste(L3, L2, L1, sep = "-"),
    x = X,
    y = Y
  ) |>
  dplyr::left_join(
    state_attr |>
      dplyr::select(row_id, abbr),
    by = "row_id"
  ) |>
  dplyr::transmute(
    state_abbr = abbr,
    x = x,
    y = y,
    group = group
  ) |>
  dplyr::left_join(
    analysis_df |>
      dplyr::select(state_abbr, rate_per_100k, law_band),
    by = "state_abbr"
  ) |>
  dplyr::filter(state_abbr != "PR")

state_centroids <- state_shapes |>
  dplyr::group_by(state_abbr) |>
  dplyr::summarize(
    cx = mean(x, na.rm = TRUE),
    cy = mean(y, na.rm = TRUE),
    rate_per_100k = dplyr::first(rate_per_100k),
    law_band = dplyr::first(law_band),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    rel_from_mean = (rate_per_100k - map_weighted_mean) / map_weighted_mean,
    # Scale up/down from centroid by relative distance from weighted mean.
    scale_factor = pmin(pmax(1 + 0.7 * rel_from_mean, 0.55), 1.55)
  )

us_center <- state_centroids |>
  dplyr::summarize(ux = mean(cx, na.rm = TRUE), uy = mean(cy, na.rm = TRUE))

exploded_map <- state_shapes |>
  dplyr::left_join(state_centroids |>
                     dplyr::select(state_abbr, cx, cy, scale_factor),
                   by = "state_abbr") |>
  dplyr::mutate(
    # Explode outward from national center to reduce overlap.
    x_shift = 0.45 * (cx - us_center$ux),
    y_shift = 0.45 * (cy - us_center$uy),
    # Original-size (unscaled) exploded border coordinates.
    x_border = x + x_shift,
    y_border = y + y_shift,
    x2 = cx + (x - cx) * scale_factor + x_shift,
    y2 = cy + (y - cy) * scale_factor + y_shift
  )

p_combo <- ggplot() +
  geom_polygon(
    data = exploded_map,
    aes(x = x2, y = y2, group = group, fill = law_band),
    color = "white",
    linewidth = 0.2
  ) +
  geom_polygon(
    data = exploded_map,
    aes(x = x_border, y = y_border, group = group),
    fill = NA,
    color = "#00e676",
    linetype = "22",
    linewidth = 0.35
  ) +
  scale_fill_manual(
    values = c(
      "F" = "#c62828",
      "D" = "#f4b6c2",
      "C" = "#d9d9d9",
      "B" = "#7fb3ff",
      "A" = "#0b3c8c"
    ),
    drop = FALSE,
    na.translate = FALSE,
    name = "Law grade band"
  ) +
  coord_equal() +
  labs(
    title = "Combined map: law strictness + mortality-distance resizing",
    subtitle = "State size changes by distance from weighted mean mortality; dashed outlines show original borders."
  ) +
  theme_void() +
  theme(legend.position = "right")

p_combo
```

## Visualizing weighted vs unweighted mean differences (6 options)

```{r mean-diff-visuals, fig.height=5.5, fig.width=10}
group_summary <- analysis_df |>
  filter(!is.na(strictness_group)) |>
  group_by(strictness_group) |>
  summarize(
    states = n(),
    mean_rate_per_100k = mean(rate_per_100k, na.rm = TRUE),
    median_rate_per_100k = median(rate_per_100k, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(strictness_group = factor(strictness_group, levels = c("Strict (4-5)", "Less strict (1-3)")))

overall_weighted_mean <- mean(analysis_df$rate_per_100k, na.rm = TRUE)
unweighted_group_mean <- mean(group_summary$mean_rate_per_100k, na.rm = TRUE)

mean_compare <- tibble::tibble(
  metric = "Deaths per 100k",
  unweighted = unweighted_group_mean,
  weighted = overall_weighted_mean
)

mean_compare

# 2) Group-size weighted bar decomposition
decomp <- group_summary |>
  arrange(strictness_group) |>
  mutate(
    xmin = dplyr::lag(cumsum(states), default = 0),
    xmax = cumsum(states)
  )

p_weighted_decomp <- ggplot(decomp) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = mean_rate_per_100k, fill = strictness_group), color = "white") +
  geom_hline(yintercept = overall_weighted_mean, linetype = "dashed", linewidth = 1, color = "black") +
  annotate(
    "text",
    x = sum(decomp$states) + 1,
    y = overall_weighted_mean + 0.35,
    label = paste0(round(overall_weighted_mean, 1), " mean"),
    hjust = 0,
    vjust = 0,
    size = 3.1
  ) +
  scale_fill_manual(
    values = c(
      "Less strict (1-3)" = "#c62828",
      "Strict (4-5)" = "#0b3c8c"
    )
  ) +
  scale_x_continuous(
    limits = c(0, sum(decomp$states) + 3),
    breaks = seq(0, sum(decomp$states), by = 10),
    minor_breaks = NULL
  ) +
  labs(
    title = "2) Weighted decomposition (bar width = group share)",
    subtitle = "Dashed line and label indicate the weighted overall mean.",
    x = "Cumulative state count",
    y = "Group mean deaths per 100k",
    fill = "Group"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())

p_weighted_decomp

# 6) Violin + box by group with mean markers
lowest_df <- analysis_df |>
  filter(!is.na(strictness_group), !is.na(rate_per_100k), !is.na(state_abbr)) |>
  group_by(strictness_group) |>
  slice_min(order_by = rate_per_100k, n = 1, with_ties = FALSE) |>
  ungroup()

median_df <- analysis_df |>
  filter(!is.na(strictness_group), !is.na(rate_per_100k), !is.na(state_abbr)) |>
  group_by(strictness_group) |>
  mutate(
    grp_median = median(rate_per_100k, na.rm = TRUE),
    med_dist = abs(rate_per_100k - grp_median)
  ) |>
  slice_min(order_by = med_dist, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(-grp_median, -med_dist)

highest_df <- analysis_df |>
  filter(!is.na(strictness_group), !is.na(rate_per_100k), !is.na(state_abbr)) |>
  group_by(strictness_group) |>
  slice_max(order_by = rate_per_100k, n = 1, with_ties = FALSE) |>
  ungroup()

violin_labels <- bind_rows(
  lowest_df |> mutate(label_type = "low"),
  median_df |> mutate(label_type = "median"),
  highest_df |> mutate(label_type = "high")
) |>
  group_by(strictness_group, label_type) |>
  slice_head(n = 1) |>
  ungroup() |>
  mutate(state_abbr = toupper(state_abbr))

p_violin <- ggplot(analysis_df, aes(x = strictness_group, y = rate_per_100k, fill = strictness_group)) +
  geom_violin(alpha = 0.35, trim = FALSE) +
  geom_boxplot(width = 0.2, outlier.shape = NA) +
  geom_point(data = violin_labels, color = "gray45", size = 2.5) +
  geom_text(
    data = violin_labels,
    aes(label = state_abbr),
    color = "gray30",
    nudge_x = 0.25,
    hjust = 0,
    vjust = 0,
    size = 3.5,
    show.legend = FALSE
  ) +
  stat_summary(fun = mean, geom = "point", shape = 21, size = 3, fill = "yellow") +
  geom_hline(yintercept = overall_weighted_mean, linetype = "dashed", linewidth = 1, color = "black") +
  labs(
    title = "6) Distribution by group with quantile box (IQR)",
    x = "Strictness group",
    y = "Deaths per 100k",
    fill = "Group"
  ) +
  scale_y_continuous(
    breaks = seq(0, ceiling(max(analysis_df$rate_per_100k, na.rm = TRUE) / 10) * 10, by = 10),
    minor_breaks = NULL
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(color = "gray80", linewidth = 0.3)
  )

p_violin

distribution_classification <- analysis_df |>
  filter(!is.na(strictness_group), !is.na(rate_per_100k)) |>
  group_by(strictness_group) |>
  summarize(
    n = n(),
    mean_rate = mean(rate_per_100k, na.rm = TRUE),
    median_rate = median(rate_per_100k, na.rm = TRUE),
    sd_rate = sd(rate_per_100k, na.rm = TRUE),
    skew = {
      m <- mean(rate_per_100k, na.rm = TRUE)
      s <- sd(rate_per_100k, na.rm = TRUE)
      if (is.na(s) || s == 0) NA_real_ else mean((rate_per_100k - m)^3, na.rm = TRUE) / (s^3)
    },
    shapiro_p = if (n() >= 3 && n() <= 5000) shapiro.test(rate_per_100k)$p.value else NA_real_,
    .groups = "drop"
  ) |>
  mutate(
    distribution_class = dplyr::case_when(
      is.na(skew) ~ "Undetermined",
      abs(skew) < 0.2 & !is.na(shapiro_p) & shapiro_p > 0.05 ~ "Approximately normal",
      skew >= 0.75 ~ "Right skew",
      skew >= 0.2 ~ "Mild right skew",
      skew <= -0.75 ~ "Left skew",
      skew <= -0.2 ~ "Mild left skew",
      TRUE ~ "Non-normal, near symmetric"
    )
  ) |>
  select(strictness_group, n, mean_rate, median_rate, skew, shapiro_p, distribution_class)

distribution_classification
```

## Correlation and regression (association, not causation)

```{r stats}
# cor.test needs at least 2 finite observations; we already sanity-check above,
# but keep a guard so knitting fails gracefully if inputs change.
if (sum(is.finite(analysis_df$likert) & is.finite(analysis_df$rate_per_100k)) < 3) {
  stop("Not enough finite observations for correlation/regression after filtering.")
}

cor_spear <- cor.test(analysis_df$likert, analysis_df$rate_per_100k, method = "spearman")
lm_fit <- lm(rate_per_100k ~ likert, data = analysis_df)

cor_spear
summary(lm_fit)
```

# PPT generation (Path B): officer + rvg

This section creates a PowerPoint file with **editable vector charts**.
After creating the PPT, open it in PowerPoint and **Ungroup** the chart objects (often twice) to animate components.

```{r make-ppt}
ppt <- read_pptx()

# Slide 1: Title
ppt <- ppt |>
  add_slide(layout = "Title Slide", master = "Office Theme") |>
  ph_with(
    "Do Stricter Gun Laws Reduce Firearm Mortality?",
    location = ph_location_type(type = "ctrTitle")
  ) |>
  ph_with(
    glue("CDC API + GIFFORDS Likert • Latest year: {analysis_year}"),
    location = ph_location_type(type = "subTitle")
  )

# Slide 2: Mortality heatmap
ppt <- ppt |>
  add_slide(layout = "Title and Content", master = "Office Theme") |>
  ph_with("Firearm mortality by state (CDC)", location = ph_location_type(type = "title")) |>
  ph_with(dml(ggobj = p_mortality), location = ph_location_type(type = "body"))

# Slide 3: Laws heatmap
ppt <- ppt |>
  add_slide(layout = "Title and Content", master = "Office Theme") |>
  ph_with("Gun-law strictness (Likert 1–5)", location = ph_location_type(type = "title")) |>
  ph_with(dml(ggobj = p_laws), location = ph_location_type(type = "body"))

# Slide 4: Association plot
p_scatter <- ggplot(analysis_df, aes(x = likert, y = rate_per_100k)) +
  geom_jitter(width = 0.2, height = 0, size = 3, alpha = 0.85) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(breaks = 1:5) +
  labs(
    title = "Association: strictness vs mortality",
    subtitle = glue("Spearman rho = {round(unname(cor_spear$estimate), 2)} (p={signif(cor_spear$p.value, 3)})"),
    x = "Strictness (1 = lax … 5 = strict)",
    y = "Deaths per 100k"
  ) +
  theme_minimal()

ppt <- ppt |>
  add_slide(layout = "Title and Content", master = "Office Theme") |>
  ph_with("Do stricter laws correlate with lower mortality?", location = ph_location_type(type = "title")) |>
  ph_with(dml(ggobj = p_scatter), location = ph_location_type(type = "body"))

# Slide 5: Methods / notes
methods_text <- paste(
  "Data:",
  "• Firearm mortality rates from CDC (data.cdc.gov) via SODA API (state + county).",
  "• Mortality time window begins 1999 for a consistent modern series.",
  "",
  "Gun-law strictness:",
  "• GIFFORDS annual state grades (A+ to F) are scraped from public scorecard pages (2012–2024).",
  "• Grades are collapsed to a 5-point Likert scale:",
  "  5=A-range, 4=B-range, 3=C-range, 2=D-range, 1=F.",
  "• Scraping safeguards: cached CSV output, light pacing between requests, and fallback parsing.",
  "",
  "Year alignment:",
  "• The analysis year is set to the most recent year available in BOTH CDC and GIFFORDS (law_year_used).",
  "",
  "Analysis:",
  "• Cross-sectional summary by Likert bin, Spearman correlation, and linear regression.",
  "• Results describe association only; confounding and policy endogeneity remain limitations.",
  sep = "\n"
)

ppt <- ppt |>
  add_slide(layout = "Title and Content", master = "Office Theme") |>
  ph_with("Methods & limitations (for your write-up)", location = ph_location_type(type = "title")) |>
  ph_with(methods_text, location = ph_location_type(type = "body"))

out_pptx <- "Gun_Laws_CDC_GIFFORDS.pptx"
print(ppt, target = out_pptx)

out_pptx
```

# How to animate in PowerPoint

1. Open the generated `.pptx`
2. Click a chart → **Right click → Group → Ungroup** (repeat if needed)
3. Now you can animate:
   - individual map layers / legends / text
   - scatter points separate from trend line
   - chart elements and labels

# Next enhancements (optional)

- Add a **panel** across years (1999–present) with a mixed model or state fixed effects
- Add outlier slides highlighting states with strict laws but high rates (and vice versa)
