---
title: "Story 3: How Much Do Data Practitioners Get Paid?"
author: "Randy Howk"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
    theme: readable
  pdf_document:
    toc: true
    number_sections: true
fontsize: 11pt
geometry: margin=1in
---

# Overview

This story answers a simple question: **How much do data practitioners get paid, and how much does that vary by role and by state?**

The assignment rubric emphasizes clarity, simplicity, saliency, and fidelity, so this draft uses a **Tufte-style low data-to-ink approach**:

- a **ranked dot plot** to show which role descriptors tend to command the highest average pay
- a **compact state-by-role heatmap** to show how geography shifts pay within each role
- direct labels, restrained color, and minimal non-data ink

# Story development focus

## Focus point: visual choice + slide messaging technique

The development focus for this story is **visual choice** supported by a clear **slide/story message**.

The design decision is intentional:

1. First answer the role question with the clearest possible visual: a **ranked dot plot**.
2. Then answer the geography question with a **dense but readable heatmap** that lets the reader scan all states at once.
3. Strip away chart junk so the viewer sees the story immediately.

That focus supports the rubric in four ways:

- **Simplicity:** no 3D effects, heavy borders, decorative legends, or redundant labels
- **Saliency:** the highest-paying roles and highest-paying states visually stand out first
- **Efficacy:** the two visuals divide the analytical task into two easy reading steps
- **Amity:** the story can be understood with very little effort

# Data and methods

## Data source

This analysis uses the **U.S. Bureau of Labor Statistics (BLS) Occupational Employment and Wage Statistics (OEWS), May 2024 state wage estimates**.

Why this source:

- it is a credible public source
- it provides state-level wage estimates
- it reports average annual wages by occupation
- it is more reliable for this assignment than crowdsourced salary sites

## Role mapping

A practical limitation is that BLS reports salary by **standard occupation**, not by every employer-specific job title.

So this first stab maps broad role descriptors to the closest BLS occupational titles:

| Role descriptor | BLS occupation used | Why |
|---|---|---|
| Data Scientist | Data Scientists | direct match |
| Data Analyst | Operations Research Analysts | close analytical/decision-support match |
| Data Engineer | Database Architects | closest published data-platform / warehouse design role |
| Data Architect | Database Architects | direct architecture-oriented data role |
| Business Analyst | Management Analysts | closest published business analysis / process improvement role |

> **Important limitation:** this mapping is an analytical approximation, not a claim that every employer uses these titles the same way. That tradeoff preserves **fidelity** by using a trustworthy wage source while still answering the assignment's broader "data practitioner" question.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(readxl)
  library(janitor)
  library(scales)
  library(glue)
  library(forcats)
  library(knitr)
})

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.major.y = element_blank(),
      axis.title = element_text(face = "bold"),
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(size = 11, margin = margin(b = 8)),
      plot.caption = element_text(size = 9, color = "grey30"),
      legend.position = "top"
    )
)
```

# Import BLS state wage data

```{r import-bls}
# May 2024 OEWS state file distributed by BLS as a zip that contains an XLSX.
bls_zip_url <- "https://www.bls.gov/oes/special-requests/oesm24st.zip"
dir.create("data", showWarnings = FALSE)
zip_path <- file.path("data", "oesm24st.zip")

if (!file.exists(zip_path)) {
  download.file(bls_zip_url, destfile = zip_path, mode = "wb")
}

unzip_dir <- file.path("data", "oesm24st")
dir.create(unzip_dir, showWarnings = FALSE)
unzip(zip_path, exdir = unzip_dir)

xlsx_file <- list.files(unzip_dir, pattern = "\\.xlsx$", full.names = TRUE)
stopifnot(length(xlsx_file) >= 1)

# Read the first worksheet and normalize names.
# The BLS state workbook layout is stable, but this code stays slightly defensive.
bls_raw <- read_excel(xlsx_file[1], sheet = 1) |> clean_names()

# Keep only the columns we need if they exist.
required_cols <- c("area_title", "occ_title", "tot_emp", "a_mean")
missing_cols <- setdiff(required_cols, names(bls_raw))
if (length(missing_cols) > 0) {
  stop(glue("Missing expected BLS columns: {paste(missing_cols, collapse = ', ')}"))
}

glimpse(bls_raw)
```

# Clean and map the occupations

```{r clean-map}
role_crosswalk <- tribble(
  ~role_descriptor,   ~occ_title,
  "Data Scientist",  "Data Scientists",
  "Data Analyst",    "Operations Research Analysts",
  "Data Engineer",   "Database Architects",
  "Data Architect",  "Database Architects",
  "Business Analyst", "Management Analysts"
)

salary_state <- bls_raw |>
  transmute(
    state = area_title,
    occ_title,
    tot_emp = suppressWarnings(as.numeric(tot_emp)),
    a_mean = suppressWarnings(as.numeric(a_mean))
  ) |>
  inner_join(role_crosswalk, by = "occ_title") |>
  filter(!is.na(a_mean)) |>
  filter(!state %in% c("United States", "Guam", "Puerto Rico", "Virgin Islands")) |>
  distinct(state, role_descriptor, .keep_all = TRUE)

# Order roles by the median state wage so the story reads from lower to higher pay.
role_order <- salary_state |>
  group_by(role_descriptor) |>
  summarise(median_state_salary = median(a_mean, na.rm = TRUE), .groups = "drop") |>
  arrange(median_state_salary) |>
  pull(role_descriptor)

state_order <- salary_state |>
  group_by(state) |>
  summarise(overall_state_median = median(a_mean, na.rm = TRUE), .groups = "drop") |>
  arrange(overall_state_median) |>
  pull(state)

salary_state <- salary_state |>
  mutate(
    role_descriptor = factor(role_descriptor, levels = role_order),
    state = factor(state, levels = state_order)
  )

summary_by_role <- salary_state |>
  group_by(role_descriptor) |>
  summarise(
    median_state_salary = median(a_mean, na.rm = TRUE),
    lowest_state_salary = min(a_mean, na.rm = TRUE),
    highest_state_salary = max(a_mean, na.rm = TRUE),
    spread = highest_state_salary - lowest_state_salary,
    .groups = "drop"
  ) |>
  arrange(median_state_salary)

summary_by_role
```

# Key takeaways

```{r key-metrics}
role_leader <- summary_by_role |> slice_max(median_state_salary, n = 1)
role_laggard <- summary_by_role |> slice_min(median_state_salary, n = 1)
widest_spread <- summary_by_role |> slice_max(spread, n = 1)

highest_observation <- salary_state |> slice_max(a_mean, n = 1)
lowest_observation  <- salary_state |> slice_min(a_mean, n = 1)
```

- The highest-paid role, using the **median across states**, is **`r role_leader$role_descriptor`** at about **`r dollar(role_leader$median_state_salary)`**.
- The lowest-paid role in this role set is **`r role_laggard$role_descriptor`** at about **`r dollar(role_laggard$median_state_salary)`**.
- The widest state-to-state pay spread appears in **`r widest_spread$role_descriptor`**, where the gap between the lowest- and highest-paying states is about **`r dollar(widest_spread$spread)`**.
- The single highest state-role estimate in this mapped set is **`r highest_observation$role_descriptor`** in **`r highest_observation$state`** at **`r dollar(highest_observation$a_mean)`**.
- The single lowest state-role estimate in this mapped set is **`r lowest_observation$role_descriptor`** in **`r lowest_observation$state`** at **`r dollar(lowest_observation$a_mean)`**.

# Visual 1: Which role descriptors pay more?

```{r role-dot-plot, fig.height=5.5}
plot_role <- summary_by_role |>
  ggplot(aes(x = median_state_salary, y = fct_reorder(role_descriptor, median_state_salary))) +
  geom_segment(
    aes(x = 0, xend = median_state_salary, yend = role_descriptor),
    linewidth = 0.7,
    color = "grey75"
  ) +
  geom_point(size = 3) +
  geom_text(
    aes(label = dollar(median_state_salary)),
    hjust = -0.1,
    size = 3.5
  ) +
  scale_x_continuous(labels = dollar_format(), expand = expansion(mult = c(0, .12))) +
  labs(
    title = "Role descriptor matters more than the label 'data practitioner'",
    subtitle = "Median state-level average salary, May 2024 BLS OEWS estimates",
    x = "Average annual salary",
    y = NULL,
    caption = "Direct labels replace a legend. Ordering is intentional: lowest to highest typical state pay."
  )

plot_role
```

## Why this chart works

This chart is the simplest way to answer the first half of the question.

- It avoids a busy bar chart.
- It ranks roles immediately.
- It uses direct labels instead of making the reader bounce between axes and legends.
- It foregrounds the **message** rather than the decoration.

# Visual 2: How much does state change pay?

```{r heatmap, fig.height=11}
plot_heatmap <- salary_state |>
  ggplot(aes(x = role_descriptor, y = state, fill = a_mean)) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_gradient(
    low = "grey95",
    high = "grey25",
    labels = dollar_format()
  ) +
  labs(
    title = "State still matters, but the pattern is easiest to see role-by-role",
    subtitle = "Each tile is a BLS state average annual wage estimate for the mapped occupation",
    x = NULL,
    y = NULL,
    fill = "Average salary",
    caption = "States are ordered by their across-role median salary. Roles are ordered by their median state salary."
  ) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(face = "bold"),
    axis.text.y = element_text(size = 7),
    legend.key.width = unit(1.4, "in")
  )

plot_heatmap
```

## Why this chart works

The heatmap answers the second half of the question efficiently.

A small-multiple map or a set of fifty-state bar charts would use much more ink and force much more effort from the reader. This heatmap keeps the task simple:

- compare **down columns** to see how state changes salary within a role
- compare **across rows** to see whether some states tend to pay more across roles
- use one compact view instead of several repetitive charts

# Supporting table

```{r supporting-table}
summary_table <- salary_state |>
  group_by(role_descriptor) |>
  summarise(
    typical_state_salary = median(a_mean, na.rm = TRUE),
    min_state_salary = min(a_mean, na.rm = TRUE),
    max_state_salary = max(a_mean, na.rm = TRUE),
    state_spread = max_state_salary - min_state_salary,
    .groups = "drop"
  ) |>
  arrange(desc(typical_state_salary)) |>
  mutate(across(where(is.numeric), dollar))

kable(summary_table, caption = "Compact summary of role-level variation across states")
```

# Interpretation

The story is not simply that "data people earn a lot." The more useful story is:

1. **Role descriptor changes expected pay materially.** The gap between the highest- and lowest-paying mapped roles is large enough that a generic label like "data practitioner" hides meaningful variation.
2. **State changes pay inside each role.** Even within the same mapped occupation, geography shifts average salary substantially.
3. **Role and state should be read together.** A person deciding between roles, locations, or recruiting markets needs both dimensions in the same story.

# Rubric alignment

## Simplicity

The visuals remove nonessential ink:

- no 3D, gradients for decoration, or chart borders
- no duplicated legends and axis labels
- no overloaded dashboard layout

## Fidelity

The analysis uses a reputable public source and states its main limitation openly: role descriptors are mapped to standard occupations rather than treated as identical.

## Utility

The charts answer the exact assignment prompt: variation in average salary by **role descriptor** and by **state**.

## Saliency

The ranked role plot highlights the main message first, while the heatmap makes the state pattern visible without distracting from the headline.

## Uniformity

Both visuals use the same visual grammar: minimalist theme, direct labeling, limited color, and ordered categories.

## Amity

The reader should understand the story quickly because the structure is sequential:

- first: which roles pay more
- second: where states shift the answer

# What I would improve in a second draft

If you want to strengthen this submission further, the next version could:

1. add a short annotation layer that calls out the top and bottom states for one or two roles
2. replace the summary table with a labeled slope/dumbbell chart for state spread by role
3. test an alternate mapping for **Data Analyst** if your instructor prefers a different BLS proxy occupation
4. add a small note on cost of living, while keeping the main chart on nominal pay so the story stays faithful to the source data

# Sources

- U.S. Bureau of Labor Statistics, Occupational Employment and Wage Statistics (OEWS), May 2024 state wage estimates
- U.S. Bureau of Labor Statistics occupation definitions for Data Scientists, Operations Research Analysts, Database Architects, and Management Analysts
